\chapter{Weekly Duties}
\section{Check for stable}
There are two opportunities to declare a stable each week. The first occurs with Thursday daily 
cumulative build, the second with Friday's weekly build.  We check Thursday's build and if it there
are problems which might prevent a stable, we try and address them with the hope that the weekly
cumulative might be "stable".

These build constitute the stable candidates.  The criteria for a stable are:
\begin{enumerate}
\item Must be a cumulative
\item All aips unit tests should pass. Tests may fail but the failure must be understood and
considered to be minor (i.e. similar to tDirectory failure on linux with NFS mounts).
\item assay().trytests() runs without failure.
\item dishlongtests() runs without failure.
\item no obvious failures in loading the help atoms
\end{enumerate}
So what does this entail?
\begin{enumerate}
\item Examine build logs for glaring problems. 
\item Look at the unit tests
\begin{enumerate}
\item{linux\_gnu}
\item{sun4sol\_gnu}
\item{sun4sol\_ntv}
\end{enumerate}
\item Run assay
\begin{enumerate}
\item{linux\_gnu}
\item{sun4sol\_gnu}
\item{sun4sol\_ntv}
\end{enumerate}
\end{enumerate}
If the build is "stable", then as aips2adm on tarzan
\begin{enumerate}
\item Send an email to aips2-naug aips2-aoc warning of the impending build
\begin{verbatim}
Greetings,

Build xx.yyy has met the stable criteria.  We will begin building the new stable @hhmm MT.
Please use weekly until the new stable build has completed.

aips2mgr

\end{verbatim}
\item as aips2adm;
\begin{verbatim}
cd /export/aips++/scripts; sh ./do\_blessing xx.yyy
\end{verbatim}
This scripts marks xx.yyy as stable; inhales the stable into /aips++/stable; and begins the stable
build for linux and solaris.  The local sgi stable build is triggered by it's own cron job.
\item Once the stable build completes, as aips2mgr on tarzan; 
\begin{verbatim}
cd; cd scripts; ./updatestables.sh 
\end{verbatim}
This will update the stable builds area on /home/tarzan5 making binaries available to users via
aupdate.
\item Send an email to aips2-sitemgr
\begin{verbatim}
Greetings,

Build xx.yyy has been declared stable.  Please update your stable releases.

aips2mgr
\end{verbatim}
\end{enumerate}

It's best to declared a stable no later than Monday afternoon, so remote sites can
inhale the latest stable Monday evening.
\section{Check the Documentation}
A rudimentary documentation check consists of 
\begin{enumerate}
\item Point your web browser at the daily/weekly docs tree,
\item Pick the reference manual, and
\item For each package, select various modules and tools.
\end{enumerate}
More thorough testing of the docs can be done with linkscan.  Linkscan will produce a series of 
reports most useful of which are missing and broken links in the docs.
