\documentstyle[11pt]{article}
\textwidth 6.25in
\textheight 9.0in
\topmargin -1cm
\oddsidemargin 0.15in
\evensidemargin 0.15in
\marginparwidth 0.5in
\newcommand{\subsubsubsection}[1]{{\em #1}}
\begin{document}
\hfill January 25, 1992
\centerline {\bf UV Data System and Processing Requirements - An Initial Distillation}
\vskip 0.3cm
\centerline {\it AIPS++ UV Data System, Calibration and Imaging Group\footnote
{Current membership: C.Flatters, M.Holdaway, R.Sault, D.L.Shone}}
\vskip 0.3cm

\section {Introduction}
This document is an inital attempt to integrate the {\it data system,
calibration and imaging} requirements for AIPS++, based on the
requirements from the individual consortium members.  However, this is
by no means a complete requirements specification for the data system;
decisions by the {\it image handling} and {\it user interface} groups
are likely to have implications for data system requirements.  In
addition there will be some overlap, in that code which is developed by
the {\it image handling} and {\it fundamental library} groups will be
required by the {\it data system, calibration and imaging} group.

This document may have to be revised to incorporate GMRT requirements.

\section {General Nature of the Data}
This section includes requirements which will affect the data system 
explicitly, or implicitly because of the nature of the observed data.

\subsection {Instrumental Requirements}
A number of requirements are generally applicable to
single dish and interferometer data:
\begin{itemize}
  \item A full measurement of the electromagnetic field requires
        four complex (polarisation) parameters, which can be converted to
        Stokes parameters; support for this should be fundamental.
  \item Polarisation measurements may be time switched if all four
        polarisation parameters are not correlated simultaneously.
  \item Multiple frequency bands may be simultaneously observed
	({\it e.g.\enspace}for observing multiple lines simultaneously or
	multi-frequency synthesis), with variable numbers of channels within
	each band.
  \item The frequency axis may be non-linear ({\it e.g.\enspace}as
	produced by acousto-optical spectrometers) and time variable.
  \item Data for combination from different observations may have different
        (but overlapping) numbers of spectral channels and channel
        widths. These may need to be accomodated within the one dataset.
  \item Focal plane arrays with arbitrary geometry ({\it e.g.\enspace}field 
        rotation through the observation) characteristics must be supported.
  \item Mosaiced observations may have many ($\sim$1000) pointing centres.
        This must be supported at both the {\it uv }dataset and image
	dataset level.
  \item The measured polarisations, frequencies and pointing centres may be
	rapidly time switched (e.g. these may change every integration).
  \item Time-series data of profiles and visibilities (e.g. pulsar data with
	bin number as a data axis) needs to be supported.
  \item Error measures or estimates ({\it e.g.\enspace}weights) should be 
        regarded as standard in an observation.
  \item Correlation data may be 16-bit integers or 32-bit floating
        point.
\end{itemize}

\subsubsection {Requirements specific to single dish observations}
\begin{itemize}
  \item Total power time series.
  \item Total power measurements, irregulary spaced in space and/or time
        for imaging.
  \item Series of spectra, including the case where the size
        of the spectrum is variable.
  \item Bit-field data for raw pulsar data.
  \item Interferometer autocorrelation data.
\end{itemize}

\subsubsection{Requirements specific to interferometry}
Interferometer arrays should be regarded as being generally inhomogenous
(although this should not preclude taking advantage of homogeneity where 
possible) in a number of ways which should be handled transparently 
whenever possible:
\begin{itemize}
  \item Polarisation characteristics ({\it e.g.\enspace}linear or circular 
        feeds and equatorial or alt-azimuth mounts) differ among antennas 
        or arrays of antennas.
  \item Antenna sizes and system temperatures may differ widely.
  \item The procedures and input data for {\it a priori\/} antenna
        calibration may vary from antenna to antenna.
  \item Space VLBI will require support for variable antenna 
        positions.
  \item Integrations times may vary from baseline to baseline.
\end{itemize}

\subsubsection{VLBI}
VLBI support is implicit in many requirements, but in addition,
support must be provided for the many correlator formats
which will prevail, including MkII, S-2 and K-4, as well as the
MkIII and VLBA modes.

The data system should support the merging of correlation data and
calibration data from different correlators and allow the user to
deal with duplicate correlations.

\subsubsection {Monitor/history data and site/instrument specific data}
There should be support for history associated with
data.  It should be possible to import and make use of history information
from ``previous incarnations'', such as on-line observing logs.
\begin{itemize}
  \item Monitor and meterological data need to go directly into dataset.
  \item Easy access to observe-time information which cannot be put into
        the dataset at the time of observation.
  \item Flexible addition to the dataset of observatory dependent monitor 
        data.
\end{itemize}

\subsubsection {Other observed data types}
It is important that the data system be as extensible as possible to support
new data data types in future:
\begin{itemize}
  \item Triple correlation data, including the case where one of the
        visibilities has a different frequency.
  \item Optical interferometer data.
  \item Images; at some level, the data system should support the 
        requirements of image handling, rather than having one data system
        for images and another for all other data.  In general, 
        ``vector images'' should be supported, allowing scope for 
        complex images, with associated errors {\it etc.}, as well
        as double precision images.
\end{itemize}

\subsection {General (Non-Instrumental) Data-System Requirements} 
\begin{itemize}
  \item The data system file format(s) should be accessible from all
        supported machines without conversion.
  \item There should be no distinction between ``multi-source'' and 
        ``single-source'' files.
  \item Applications should function on data and data-cubes in
        arbitrary sort order; all sorting should be hidden to
        the user.
  \item Coordinate handling must be very general:
	\begin{itemize}
	  \item Support for data with non-regular increments in coordinate
	        axes (such as an optical velocity axis).
 	  \item	Coordinate systems and ephemeris information required
	        for astrometry, and near field imaging will be supported.
	\end{itemize}
        This will also assist in the merging of data sets.
  \item Support for data errors should be fundamental to the data system,
        providing a basis for support of data error handling in a
        variety of applications:
        \begin{itemize}
          \item Easy generation of error models;
          \item Error propagation through a series of tasks;
          \item Error images associated with astronomical images;
          \item Plotting error bars on spectra and profiles;
          \item Automatic warning if contour levels are below noise;
          \item Error-based blanking in display of results;
          \item Properly formatted errors when data are extracted for 
                tabulation.
        \end{itemize}
  \item Simultaneous processing of ``associated'' datasets, such as
        different ({\it e.g.\enspace}by calibration, integration time,
        fringe fitting {\it etc.}) versions of the same observation, so that
        the best can be selected later.
  \item A processing history should be maintained for each data set that
        can be reviewed by the astronomer.
  \item The user should have access to both data values and the `header'
        values that govern the interpretation of the data.
\end{itemize}

\section {Data Display and Editing}
Users must be able to display and interact with data in ways which are
much more flexible and effective than in AIPS.  This has long been regarded
as vital to system with few baslines, such as VLBI, but is also likely to prove
useful in processing other forms of data.  Data display and editing should be
seen as generic tools, applicable to single dish and other data as well as
interferometer visibilities.  To this end, some degree of data visualisation
should perhaps be seen as an integral, or at least closely coupled part of the
data system.

Many ``viewing strategies'' are desirable for display and interactive 
flagging of data:
\begin{itemize}
  \item Baseline by baseline display and editing (including multiple, 
        simultaneous baselines).
  \item Display of data aggregated in various ways ({\it e.g.\enspace}averaged
        over a number of spectral channels).
  \item Interactive selection of data to be displayed:
        \begin{itemize}
          \item taking arbitrary cuts ({\it e.g.\enspace}circular, radial or
                a user-defined locus) through the {\it uv}-plane;
          \item setting windows in space and/or time;
          \item expanding aggregates ({\it e.g.\enspace}pointing and
                clicking on an averaged multi-channel visibility to show 
                the component spectrum).
        \end{itemize}
  \item Display of generic model data ({\it i.e. }from CLEAN components,
        gaussian fitting {\it etc.}) with observed and/or processed data, 
        and display of data with model subtracted or divided.
  \item Flagging should be reversible, with the ability to store flagging
        information and apply this on the fly.
  \item Flagging on the basis of monitor/observing log data.
  \item Flagging from ``consistency check'' information, in particular
        redundancy, where possible, or crossing-points in the {\it uv}-plane.
  \item ``Intelligent'' automated flagging for large datasets.
\end{itemize}


\section {Calibration and Associated Processing}
\begin{itemize}
  \item Calibration, like flagging, should be reversible, with
        the ability to store calibration information and apply it
        ``on-the-fly''.  However, it should also be possible to 
        apply calibration information ``once-and-for-all'', creating 
        a new, calibrated data-set.
  \item Calibration should be made as generic as possible;
        site-specific/instrument dependent code should be kept
        to a minimum.
  \item Calibration of data should be possible from derived tables
        of instrumental parameters such as system temperature and 
        gain {\it vs. }elevation.   It should be possible to derive 
        such tables from calibration observations.
  \item The calibration process should include flexible interpolation and
        averaging of calibration data under the control of the user.
  \item Redundancy (possibly including crossing points) should be
        used whenever possible, as an additional constraint on
        calibration and self-calibration.
  \item Model fitting should be possible in both the image and 
        {\it uv-}planes, and it should be possible to use the resultant
        model in the same way as a CLEAN component model in calibration
        and self-calibration.
  \item Fringe fitting (normally only for VLBI data) should be possible by 
        baseline, as well as globally.
  \item Cross-calibration of different instruments should be possible,
        ({\it e.g.\enspace}flux scale, pointing) particularly where data from
        different arrays are to be combined.
\end{itemize}

\section {Synthesis Imaging}
The title of this section might be construed as covering simply the
act of transforming edited, calibrated data and deconvolving the
images.  However, we wish to emphasise that the process should be
viewed as a whole, and whilst data editing, for example, is discussed
separately (because it is a generic operation, applicable
beyond interferometry), it must be possible to integrate
(self-)calibration, data inspection/editing, transformation and
deconvolution more closely than is possible in AIPS.  

It should be possible to easily ``mix-and-match'' self-calibration,
transformation, and deconvolution tools, for example, using CLEAN to
deconvolve in the early stages, and Maximum Entropy later on when
CLEAN would begin to break-down.  This also demonstrates the need to
make self-calibration use a generic model, which may be
CLEAN-components, an image, or a gaussian model.

\section{User Access to UV Data}
There need to be flexible and simple ways for the users to manipulate
their uv datasets. For example:
\begin{itemize}
\item The user should be capable of using the host systems
 capabilities and utilities to manage his datasets. The datasets should
 use a normal file name, and the user should be able to use the host
 directory hierarchy to best organise the data.
\item Tasks should generally take multiple input uv datasets
 where this makes sense. For example, the map making program should be
 capable of taking multiple input datasets, all of which contribute to the
 output images.
\item There needs to be a flexible way for the user to select
 the particular subset of data, in a dataset, to be processed. As well
 as selection based on time, antenna number, frequency, etc, it should
 be possible to select based on the values of other data (including monitor
 data). 
\end{itemize}
There need to be methods for general users to perform unusual or
different processing and analysis on their uv data. In particular:
\begin{itemize}
\item It should be possible to extract a subset of data from
 a dataset, manipulate it in some powerful (e.g. IDL-like) command language.
 This would include displaying the data and optionally
 replacing it in the dataset.
\item It should be possible to extract a subset of data from
 a dataset in a variety of formats (eg. FITS or plain text) in order
 to transfer the data to other programs or packages.  It should also
 be possible to read the modified data back into AIPS++ using the
 same formats.
\item It should be possible to import general astronomical data (eg.
      catalogues) into AIPS++.
\item For applications where the built-in tasks and command
 language features are insufficient, there needs to a program interface to allow
 the casual programmer reasonable access to the data. Some flexibility and
 efficiency can be sacrificed in making this interface comparatively simple.
 FORTRAN programmers should be supported.
\end{itemize}

\end {document}

