\documentstyle{article}
\begin{document}
\newcommand{\aipspp}{AIPS++}
\newcommand{\tbt}[4]{\left(\begin{array}{cc}#1 & #2\\ #3 & #4 
   \end{array}\right)}
\newcommand{\x}{{\bf x}}
\newcommand{\ur}{\underline{r}}
\newcommand{\urho}{\underline{\rho}}
\newcommand{\skyI}{\vec{\cal I}}
\newcommand{\visantterm}[1]
   {{#1}_i}
\newcommand{\skyantterm}[1]
   {{#1}_i\left(\urho_k\right)}
\newcommand{\visterm}[1]
   {\left[{{#1}_i\otimes{#1}^*_j}\right]}
\newcommand{\skyterm}[1]
   {\left[{{#1}_i  {\left(\urho\right)} \otimes
           {#1}^*_j{\left(\urho\right)}}\right]}
\newcommand{\skykterm}[1]
   {\left[{{#1}_i  {\left(\urho_k\right)} \otimes
           {#1}^*_j{\left(\urho_k\right)}}\right]}
\newcommand{\fpd}[2]
   {{\partial #1\over\partial #2}}
\newcommand{\spd}[3]
   {{\partial^2 #1\over\partial #2\partial #3}}
\newcommand{\fg}[2]
   {{F_{#1}G_{#2}}}
\newcommand{\ggg}[2]
   {{G_{i,#1}G^*_{j,#2}}}
\newcommand{\fgh}[3]
   {{\fpd{F_{#1}G_{#2}}{H_{#3}}}}
\newcommand{\vecterm}[1]
{\left(
   \begin{array}{c}
      {#1}_{\rm pp}\\
      {#1}_{\rm pq}\\
      {#1}_{\rm qp}\\
      {#1}_{\rm qq}
   \end{array}
\right)_{ij}}
\newcommand{\Svec}
{\left(
   \begin{array}{c}
   I\\
   Q\\
   U\\
   V\\
   \end{array}
\right)}
\newcommand{\onevec}
{\left(
   \begin{array}{c}
   1\\
   1\\
   1\\
   1\\
   \end{array}
\right)}

\title{The Generic Interferometer: I Overview of Calibration and Imaging\\
        AIPS++ Implementation Note 183}

\author{T.J. Cornwell, NRAO}
\date{1995, August 22}
\maketitle

\tableofcontents

\section{Introduction}

In a beautiful paper, Hamaker, Bregman and Sault (1995) described a new
formalism for understanding and analysing the measurement of polarized
radiation by radio-interferometers. Although the HBS paper was
restricted to a point source, Bregman later developed the formalism for
extended objects. Noordam (1995) pointed out that this leads to a
measurement equation for a Generic Interferometer (GI), that provides a
means of unifying the polarization calibration and imaging schemes for
many different synthesis arrays. 

In previous notes, I described mathematical approaches to imaging and
calibration based upon a number of simple versions of the measurement
equation (Cornwell, 1992a, 1992b). In this document, I will revisit the
analysis performed in those two documents in the light of the new
measurement equation.

I will avoid statements about design of software based upon these
equations. Instead, analysis and design will be deferred to a later
document. Hence I will try to separate concepts that are distinct,
a discipline that will be useful later during the design phase.

The purpose of this document is to consider in more detail than HBS
how calibration and imaging is to be performed for the Generic
Interferometer. The formalism is very general and the equations seem
somewhat different from those we have seen previously in
interferometry.  I will show that in fact straightforward
generalizations of existing methods of calibration and imaging are
possible. In calibration, one can derive update equations that
resemble those usually seen. In imaging, the usual concepts of dirty
image, residual image, and PSF can be similarly extended. I will show
that the general formulas thus derived can be simplified to special
cases to yield familiar results.

\section{The Measurement Equation for a Generic Interferometer}

I will restate the HBS measurement equation as described by Noordam
but with a few additions and simplifications to clarify the physical
content of the equation. In addition, I wish to correct what I believe
to be mistakes and omissions in his formulation.

A Generic Interferometer measures cross correlations between two
channels per feed, leading to a four-term cross-correlation that we
write as a vector, $\vec{V}$, subscripted by the pair of feeds
$i,j$:

\begin{equation}
\label{eqn:vterm}
\vec{V}_{\rm ij} = \vecterm{V}
\end{equation}

This basic measureable can be decomposed as follows. The measured
cross-correlation $\vec{V}$ is given by:

\begin{equation}
\label{eqn:veqn}
\vec{V}_{\rm ij} = \visterm{G} \visterm{D} \visterm{C} \vec{\cal V}_{\rm ij}
\end{equation}

\noindent where $\visantterm{G}$, $\visantterm{D}$, $\visantterm{C}$ 
are 2 by 2 matrices representing specific feed-based effects. The
operator $\otimes$ represents a direct matrix product yielding in this
case 4 by 4 matrices. $\visantterm{G}$ represents the complex gain of
the $i$'th feed, $\visantterm{D}$ the leakage, and $\visantterm{C}$ is
a fixed matrix representing the nominal feed configuration. Noordam
gives examples of these terms.  The terms $\visterm{G}$,
$\visterm{D}$, and $\visterm{C}$ are 4 by 4 matrices, and thus the
formulation is 4-dimensional. Physically the matrices
$\visantterm{G}$, $\visantterm{D}$, $\visantterm{C}$ represent
coupling between the two polarization states for each feed and so the
terms $\visterm{G}$, $\visterm{D}$, $\visterm{C}$ represent coupling
between the four polarization correlation states.  In appendix A, I
give a catalog of various forms for the calibration matrices.

A hint on reading these and succeeding equations: to get the
more usual equations, just convert matrices to scalars, ignore the
transposes, and convert the direct matrix product to a simple
product. The equations derived below should then look quite familiar.

The vector $\vec{\cal V}$ represents the visibility that would be
measured in the absence of the visibility-domain calibration effects. 
\footnote{In Noordam's note, $\vec{\cal V}$ is called $\Sigma$ and is
mistakenly called a matrix. It actually is a visibility vector.}

\begin{equation}
\label{eqn:Vterm}
\vec{\cal V}_{\rm ij} = \vecterm{\cal V}
\end{equation}

This includes only image-domain calibration effects. Thus, the feed-based 
effects must now be a function of direction, $\underline{\rho}$.

\begin{eqnarray}
\label{eqn:vveqn}
\vec{\cal V}_{\rm ij} 
    =  \sum_k & \skykterm{E}  \skykterm{P} \skykterm{F} \nonumber \\
   & S \ \skyI_k \ e^{-2\pi i \left({\ur_i-\ur_j}\right)\urho_k} 
\end{eqnarray}

\noindent where $\skyantterm{E}$ represents the feed voltage
receptivity pattern of the $i$'th feed (Noordam used $B$ for this
term), $\skyantterm{P}$ parallactic angle rotation, and
$\skyantterm{F}$ atmospheric terms including tropospheric and
ionospheric phase (Noordam used this term for Faraday rotation alone; here I
note that Faraday rotation can also be accomodated in $F$).
The term $\skyI$ represents the polarized sky brightness. In the
Stokes representation, this is given by:

\begin{equation}
\label{eqn:skyI}
\skyI=\Svec
\end{equation}

The matrix $S$ converts from the representation used in $\skyI$ to
that most naturally used in describing the interferometer.  HBS
use linear polarization as the canonical representation, for which the
$S$-matrix is:

\begin{equation}
\label{eqn:slin}
S_{\rm linear} = {1\over 2}
       \left({
       \begin{array}{cccc}
          1&1&0&0\\
          0&0&1&i\\
          0&0&1&-i\\
          1&-1&0&0
       \end{array}
       }\right)
\end{equation}

It is worth emphasizing that this is an arbitary choice and that one could,
instead, use circular polarization, in which case the $S$-matrix would be:

\begin{equation}
\label{eqn:scirc}
S_{\rm circular} = {1\over 2}
       \left({
       \begin{array}{cccc}
          1&0&0&1\\
          0&1&i&0\\
          0&1&-i&0\\
          1&0&0&-1
       \end{array}
       }\right)
\end{equation}

It may seem that $\visantterm{C}$ and $S$ are redundant since one can
choose $S$ so that $\visantterm{C}$ is a unit matrix. However, I
follow HBS and choose $S$ to be the canonical $S^{lin}$. Consequently,
$\visantterm{C}$ must be allowed to vary to suit the actual
measurement scheme of the feed.  This flexibility is required for a
system of interferometers in which a mixture of linear and circular
polarization is measured. Note that the calibration matrices will then
inevitably be more complicated than would be the case if a single,
natural representation was used.

To emphasize the true glory of the full measurement equation, here I
give the whole expression:

\begin{eqnarray}
\label{eqn:veqndiscrete}
\vec{V}_{\rm ij} 
    = & \visterm{G} \visterm{D} \visterm{C} \nonumber \\
      & \sum_k \skykterm{E}  \skykterm{P}\skykterm{F} \nonumber \\
      & S \ \skyI_k \ e^{-2\pi i \left({\ur_i-\ur_j}\right)\urho_k} 
\end{eqnarray}

It is worth making a number of comments about this very general form of 
the measurement equation:

\begin{description}
\item[Ordering of terms] The ordering of terms follows the signal
path, reading right to left in the equation.
\item[Linearity] The ME is linear in the sky brightness $\skyI$.
It is almost always non-linear in various calibration parameters.
This will have implications for solvers, as I discuss below.
\item[Time and Frequency] The formalism as presented here ignores indexing 
by time and frequency. Both of these are trivial to add but obscure
the notation.  Averaging over both time and frequency is then
accomodated easily.
\item[No new physics] There is no new physics in this formulation.
The equations are totally equivalent to those derived earlier
(see {\em e.g.} Schwab, 1984). However, I expect that the 
formulation will suggest new forms of calibration and imaging.
\item[Use of these equations] In general, one will want to
determine one or both of the calibration parameters and the sky
brightness. To determine the calibration, one will fix the sky brightness
model, and solve for the calibration parameters, perhaps by a least squares
approach. The determine the sky brightness, one will want to fix the
calibration parameters and solve for the sky brightness, using 
a deconvolution algorithm. One could alternate between calibration and
sky brightness estimation, as is done in most self-calibration 
procedures, or one could do a joint solution by some very powerful
least squares type algorithm.
\item[Analytical forms for $\skyI$] In the calibration cycle, one will
have to perform an integration over $\urho$.
This will in general be quite difficult and usually, the
integration will have to be performed numerically. The ``art'' of
programming this equation will principally be in finding quick methods
of numerical integration using, for example, FFTs.
\item[Non-invertibility] The ME is clearly not invertible for either the
sky brightness or the calibration matrices. This is really
a truism since it represents only one sample.
\item[Self-cal assumption] The fundamental assumption yielding the direct
products is that all these instrumental effects factorize per
feed. It is hard to think of any exceptions to this, apart from
problems with the correlation process itself. Note that failure of
closure due to different bandpasses at the different feeds is
actually accomodated by the summation over frequency.
\item[Non-isoplanatism] The location of the atmospheric phase terms is
actually ambiguous in most cases. Physically it belongs with the $F$ term.
In isoplanatic conditions, it cannot be distinguished easily from
the electronic term $G$, and so it is usually written as belonging in
$G$. However, if the atmosphere is non-isoplanatic then the atmospheric
phase must be included in the $F$-term. $F$ then actually decomposes a
follows:

\begin{equation}
\label{eqn:Fdef}
\skyantterm{F} = \skyantterm{F}^{troposphere} \skyantterm{F}^{ionosphere} 
\end{equation}

\item[Assumption of perfect correlation] Note that the correlation process
itself introduces errors that must be corrected, such as the van Vleck
correction. These we ignore for the moment.
\item[Completeness of information] It is assumed that the full 4-dimensional
visibility is measured. In many cases, this is not so and, for example,
only the parallel hands will be correlated. Full correction is not
then possible but one expects that good approximations will be obtained
in many cases simply by inserting zeroes in the appropriate 4-vectors.
A similar but distinct difficulty arises in those interferometers that
measure different hands at different times. Such wrinkles can be dealt
with straightforwardly but further discussion is deferred to a 
specifications document.
\end{description}

\section{Calibration of the visibility}

Calibration involves solving for the possibly unknown feed-based
matrices such as $\visantterm{G}$, $\visantterm{D}$ from calibration
observations\footnote{Note that the matrix $\visantterm{C}$ denotes
the nominal feed configuration and thus does not need to be estimated}.
Imaging means estimating $\skyI$ from data corrected for
the calibration matrices. Note that parametrization of the free
variables is probably wise. For calibration, this means that, for
example, $\visantterm{G}$ is diagonal with parameters $a_p$, $a_q$,
$\phi_p$ and $\phi_q$ describing the amplitude and phase of the gain
in the two polarization channels. For imaging, the parameters could be
the pixels of an image or the parameters of a number of Gaussian
components.

If the image parameters are known for some subset of the measurements,
then one can solve for the calibration parameters and thus
calibrate. Then from another subset one can solve for image parameters
of another object.

Suppose that we have a number of measurements $\vec{V}$ of the
visibility function, and that furthermore we have estimates of the
visibility function $\vec{V}$ that would be measured in the absence of
visibility-based calibration effects. We can then estimate the
calibration matrices by a least squares fit whereby the matrices are
adjusted to fit the data in a least-squares sense. Define an error
norm $\chi^2$:

\begin{equation}
\label{eqn:Sdef}
{\chi^2} = \sum_{ij} 
  \Delta \vec{V}^{*T}_{\rm ij} W_{ij} \Delta \vec{V}_{\rm ij}
\end{equation}

\noindent where the residual is given by:

\begin{equation}
\label{eqn:dvdef}
\Delta \vec{V} = \vec{V} - \widehat{\vec{V}}
\end{equation}

\noindent and $\widehat{\vec{V}}$ denotes an estimate:

\begin{equation}
\label{eqn:vcorr}
\widehat{\vec{V}} = \visterm{G} \visterm{D} \visterm{C} \widehat{\vec{\cal V}}_{\rm ij}
\end{equation}

\noindent and $W_{ij}$ is a (4 by 4) weight matrix. For natural weighting, 
$W$ will usually be diagonal with elements given by the inverse of the
corresponding variance. However, it is more strictly the inverse of
the covariance matrix of the errors. For uniform weighting, the usual
correction for the local density of samples will be applied.

Choosing the calibration matrices that minimize ${\chi^2}$ is a
non-linear least squares problem. To solve it, we will need the first
and second derivatives of ${\chi^2}$ with respect to the calibration
matrices. Since this gets quite complicated, I've deferred complete
exposition to another memo.  For the moment, I say only that to check
these formulas, I generated the scalar update equation for the case
where only the gain matrix need be calculated.  I do indeed obtain the
update formula used in the usual feed-based gain solution algorithms.

This is all good as far as it goes, but we probably will prefer to parametrize
the calibration matrices by a small number of parameters. We will then require
derivatives with respect to these parameters, instead of the elements of the
matrices directly. By applying the chain rule, this is quite straightforward,
if tedious.

Having solved for the calibration matrices, we can now correct for them:

\begin{equation}
\label{eqn:veqninv}
\vec{{\cal V}}_{\rm ij} = \visterm{C}^{-1} \visterm{D}^{-1} \visterm{G}^{-1} 
\vec{V}_{\rm ij}
\end{equation}

The other calibration matrices, $\skyantterm{E}$, $\skyantterm{P}$,
$\skyantterm{F}$, cannot be corrected in the visibility
domain. Instead, it is necessary to make an image. This I describe
next.

\section{The Generalized Dirty Image}

Imaging from the measurement equation requires some careful
analysis. We have to accept that there is no inverse to the
measurement equation. The {\em dirty} image is simply one means of
estimating the sky brightness from the measurements. It is convenient
because it is linear in the measurements. However, giving it special
status obscures the underlying simplicity of imaging.

The dirty image is known to be useful in imaging from simple
interferometric observations. Here I will shown how to generalize it
to this more complicated measurement equation. 

We vary the estimate of the sky brightness $\skyI_k$ to minimize the
function ${\chi^2}$. For this we need to evaluate
$\fpd{\chi^2}{\skyI_k}$. Old hands at this sort of stuff will
recognize that it looks like:

\begin{eqnarray}
\label{eqn:dSdI}
{\fpd{\chi^2}{\skyI_k}}
    = & - 2\ \Re \sum_{ij} S^{*T} \nonumber \\
      & \skykterm{F}^{*T} \skykterm{P}^{*T} \skykterm{E}^{*T} \nonumber \\
      & \visterm{C}^{*T} \visterm{D}^{*T} \visterm{G}^{*T}  \nonumber \\
      & W_{\rm ij} \ \Delta \vec{V}_{\rm ij} 
      \ e^{2\pi i \left({\ur_i-\ur_j}\right)\urho_k} 
\end{eqnarray}

It is still a long way from this to the dirty image. To make a start,
set the initial model to be zero. $\Delta\vec{V}$ then becomes
$\vec{V}$. We can read the equation from the inside out as:
\begin{itemize}
\item Apply the transposed complex conjugate of the gain $\visterm{G}$,
\item Apply the transposed complex conjugate of the D-terms $\visterm{D}$,
\item Apply the transposed complex conjugate of the feed configuration $\visterm{C}$,
\item Apply the transposed complex conjugate of the primary beam $\skykterm{E}$,
\item Apply the transposed complex conjugate of the parallactic angle $\skykterm{P}$,
\item Apply the transposed complex conjugate of the atmospheric phase term $\skykterm{F}$,
\item Apply the transposed complex conjugate of the conversion from Stokes representation,
\item Apply the inverse Fourier transform for this sample,
\item Sum over all visibility samples with complex weights, {\em i.e.\ } the
Fourier Transform for one sample.
\end{itemize}

The presence of all the transposes is a clue that we are still not
quite there. For unitary matrices, the transpose of the complex
conjugate is the inverse, and so we can recognize the correction of
the various effects.  However, there remains a normalization problem.
Furthermore, for non-unitary matrices, we have also to correct for the
amplitude of the various matrices. These points are connected to the
fact that the gradient is not a good search direction for a least
squares method. It is better to pre-multiply by the inverse of the
Hessian. Remember that the Hessian has elements: $\spd{\chi^2}{\skyI_k}{\skyI_l^T}$
Since, in general, this is not practicable we make do with
the inverse of the diagonal elements: $\spd{\chi^2}{\skyI_k}{\skyI_k^T}$
With a bit more heavy lifting, we get that it must be:

\begin{eqnarray}
\label{eqn:ddSddI}
\spd{\chi^2}{\skyI_k}{\skyI_k^T}
    = & 2\ \Re \sum_{ij} S^{*T} \nonumber \\
      & \skykterm{F}^{*T} \skykterm{P}^{*T} \skykterm{E}^{*T} \nonumber \\
      & \visterm{C}^{*T} \visterm{D}^{*T} \visterm{G}^{*T} \nonumber \\
      & {W_{\rm ij}} \ \nonumber \\
      & \visterm{G} \visterm{D} \visterm{C} \nonumber \\
      & \skykterm{E} \skykterm{P} \skykterm{F}\nonumber \\
      & S
\end{eqnarray}

An appropriate generalization of the dirty image can be defined as the
initial update direction for a least squares fit. Applying the
Newton-Raphson approach with the approximation that only the diagonal
terms are retained, we therefore have that:

\begin{equation}
\label{eqn:IDdef}
\skyI^D_k = - \left[\spd{\chi^2}{\skyI_k}{\skyI^T_k}\right]^{-1}
   {\fpd{\chi^2}{\skyI_k}}\mid_{\skyI_k=0}
\end{equation}

Note that the first term on the RHS of this equation is the inverse of
a 4 by 4 matrix. By inverting this matrix, we are correcting for the
coupling of different polarizations in the interferometer. By ignoring
the non-diagonal terms of the Hessian, we are ignoring the coupling
between different pixels in the final image. This is reasonable since,
first, the coupling is singular, and, second, it is the role of a
deconvolution algorithm or Image Solver (see below) to correct for
this coupling.

A residual image can be similarly defined as the update direction for
a given estimate of the sky brightness, $\skyI$.

Again, we should check that these equations give the right answers for
known, simple cases. Let us consider imaging of {\bf Stokes parameter
I}. The generalized dirty image, as defined in equation
\ref{eqn:IDdef}, is then simply the familiar dirty image:

\begin{equation}
I^D_k = {
   \sum_{ij} W_{\rm ij} \ \Re{\left( V_{\rm ij} 
   \ e^{2\pi i \left({\ur_i-\ur_j}\right)\urho_k} \right) }
   \over 
   \sum_{ij} {W_{\rm ij}}
   } 
\end{equation}

\noindent where $I$, $W$ and $V$ are now all scalars.

The above-mentioned approximation of dividing by the inverse of the
diagonal elements of the Hessian thus corresponds in this case to
normalizing by the summed weights, something that is common in
constructing a dirty image. This is a good approximation for a point
source\footnote{I note that for an extended source, one might wish  to
use a different approximation for the inverse Hessian} and so the
strength of an isolated point source in the generalized dirty image is
correctly represented.

Let us turn now to a case for which the full vector formulation is
needed: for {\bf full polarized observations with gain
errors}, the generalized dirty image is:

\begin{eqnarray}
\label{eqn:IDdefpol}
\skyI^D_k = 
    &\left[\sum_{ij} \Re{\left( S^{*T} \visterm{G}^{*T} {W_{\rm ij}} 
                  \ \visterm{G} S \right)}\right]^{-1}& \nonumber \\
    &\sum_{ij} \Re{\left( S^{*T} \visterm{G}^{*T} W_{\rm ij} \ \vec{V}_{\rm ij}
    \ e^{2\pi i \left({\ur_i-\ur_j}\right)\urho_k} \right)}
\end{eqnarray}

This looks a little more curious and unexpected: the formalism
requires that one corrects the gains only on average, not per sample
as is usually done, and as arose in the previous section. Furthermore,
the correction for the coupling between the different polarization
channels is performed by a matrix inversion and multiplication in the
image plane. A little thought tells one that this is actually the
correct prescription so, as Jan Noordam says in his memo, our
calibration procedures do benefit from some careful scrutiny.

Finally, let us look at the case of {\bf primary beam correction}, and
its big brother, mosaicing. Here the vector formulation yields:

\begin{eqnarray}
\label{eqn:IDdefpb}
\skyI^D_k = 
&\left[\sum_{ij} \Re{\left(S^{*T} \skykterm{E}^{*T} W_{\rm ij} 
      \ \skykterm{E} S \right)}\right] ^{-1} &\nonumber \\
    &\sum_{ij} \Re{\left(S^{*T} \skykterm{E}^{*T} W_{\rm ij} \ \vec{V}_{\rm ij} 
     \ e^{2\pi i \left({\ur_i-\ur_j}\right)\urho_k} \right)}
\end{eqnarray}

Thus again, the primary beam division is done only on
average. However, unlike the case of gain errors, this result is
somewhat familiar from mosaicing theory (see {\em e.g.\ } Cornwell,
Holdaway, and Uson, 1993), except that it is a vector formulation. In
this case, one has a linear method that corrects for an image plane
effect in an intuitive way.

In summary, I have shown that the GI measurement equation can be used
to define a generalized Dirty Image that seems to be quite useful in
representing the sky brightness at some level of approximation. Note
that the so-called image plane effects, the primary beam, the
parallactic angle and the atmospheric phase effects, can be corrected
without going through a full non-linear deconvolution. Another way
of understanding this evidently unexpected result is that the dirty
image is best viewed as the result of one form of deconvolution, special
only in its linearity. This same point is made in my memo on the
A-matrix formalism (Cornwell, 1992a) but is presented more explicitly
here.

\section{Image and Calibration Solvers}

The sky brightness and the calibration matrices now appear in the
formalism on an equal basis. This is in accord with our experience
that one can use interferometers to image the sky or to calibrate the
interferometer.

Deconvolution can now be described as a processing of solving for the
sky brightness $\skyI$. It is probably better to use a term such as
{\em Image Solver} since the connection to classic deconvolution is
getting harder and harder to follow.  Image Solvers such as CLEAN,
including all variants such as the Clark and Schwab-Cotton algorithms,
and MEM, can be written in terms of ${\chi^2}$ and the gradient terms
described above. Thus the machinery for deconvolution can be separated
from the machinery for the measurement equation. This is the essence
of abstraction.

Solving for calibration is conceptually much more straightforward than
imaging. Let us assume that we have a model for the sky brightness,
$\widehat{\skyI}$. We then find those calibration matrices that minimize
$\chi^2$. I described above, how one might imagine doing this by a
simple gradient search method analogous to that using in imaging. One
caution is in order: since the calibration matrices are parametrized by
other free variables and since such parametrizations are likely to be
strongly non-linear, it is likely that a simple approach like that
advocated for imaging will be ineffective and so a more complicated
algorithm may be required. Even if this is true, it is probably also
true that all that is required from the measurement equation is
knowledge of first and second derivatives.

\section{Open and unresolved questions}

\begin{enumerate}
\item Can the coherence matrix be used as the prime observable instead
of the coherence vector? The mixed product rule cannot be applied
directly but the connection should be possible. Does it have any
advantages? One advantage is that the eigenvalues of the coherence
matrix are the two independent states of the radiation field and
so, for example, one obtains the entropy of a polarized radiation
field by summing the entropies of the two eigenvalues. Presumably
something similar must apply to CLEAN. This question, therefore,
is most important when considering Image Solvers.
\item Can this formalism be usefully applied to single dish observations?
For mosaicing, this would be convenient. One would need to add terms
for total power offsets that are neglected in most synthesis
applications.
\item Does this formalism lend itself naturally to the expression of
fringe fitting algorithms? The answer is almost certainly yes.
\end{enumerate}

\section{Summary}

The GI measurement equation lends itself to the following:

\begin{itemize}
\item Conceptually simple (but mathematically elaborate) calibration 
algorithms based upon derivatives of the misfit, $\chi^2$, with 
respect to the calibration matrices.
\item A generalized dirty image in which image plane effects can be
corrected to some degree,
\item Image Solvers that interface to the measurement equation solely
through derivatives of the misfit, $\chi^2$.
\end{itemize}

In all cases, the general equations can be simplified to yield familiar
results, thus reassuring us of the basic correctness of this approach.

\section*{Acknowledgements}

Fred Schwab provided invaluable advice and references on Matrix Calculus
and calibration.

\section*{References}

\noindent Cornwell, T.J., 1992a, AIPS++ Implementation Note 147.

\noindent Cornwell, T.J., 1992b, AIPS++ Implementation Note 148.

\noindent Cornwell, T.J., Holdaway, M.H., and Uson, J.M., 1993, 
Astron. \& Astrophys. {\bf 271}, 697-713.

\noindent Hamaker, J.P., Bregman, J.D., and Sault, R.J., 1995,
{\em Understanding radio polarimetry: I Mathematical foundations},
submitted to A\&A.

\noindent Noordam, J., 1995, AIPS++ Implementation Note 182.

\noindent Schwab, F.R., 1984, {\em Comment on the Use of Linearly Polarized
Feeds for the VLBA}, VLBA memo 337.

\section*{Appendix A: Catalog of Matrices}

Here I repeat Noordam's description of the various matrices:

\begin{itemize}
\item $G_{i}$ represents complex gain (amplitude, phase) per
IF channel. It includes the effects of all the electronics after the
feed (amplifiers, mixers, LO, cables etc), excluding the correlator of
course. 

\begin{equation}
  G_{i} 
  = 
  \tbt{g_{i\rm p}} {0} {0} {g_{i\rm q}}
  \label{eq:Gi}
\end{equation}

\item ${D}_{i}$ represents the signal `leakage' between the two
receptors in a feed, due to deviations from nominal. The real part
corresponds roughly to a dipole alignment error, and the imaginary
part to ellipticity. 

\begin{equation}
  {D}_{i} 
  = 
  \tbt{1} {d_{i\rm p}} {-d_{i\rm q}} {1}
  \label{eq:Di}
\end{equation}

{\it NB: Other processes may also contribute to leakage.  For example,
cross-talk between the two IF-channels can be described by adding a
non-zero mutual coupling factor $c_{i}$ to the off-diagonal terms}.
\item ${C}_{i}$ represents the nominal feed configuration, which
includes rotation of the dipoles w.r.t. the antenna (e.g. WSRT) , and
conversion from linear to circular polarisation (e.g. VLA).
\item ${E}_{i}$ represents the feed beam response.  For the
important case of a dominating compact source in the centre of the
field (and an axially symmetric antenna-feed combination), it is the
unit matrix.
\item ${P}_{i}$ represents parallactic angle $\phi$,
i.e. rotation of the antenna w.r.t. the sky.  For an {\it equatorial}
antenna like used in the WSRT, it is a unit matrix. For an {\it alt-az}
antenna, $\phi$ varies smoothly with Hour Angle (HA), as a function of
LATitude and DEClination:
\begin{equation}
  {P}_{i}^{alt-az} 
  = 
  \tbt{\cos\phi(t)} {\sin\phi(t)} {-\sin\phi(t)} {\cos\phi(t)}  
  \label{eq:Pi-alt-az}
\end{equation}
\item ${F}_{i}$ represents (ionospheric) Faraday rotation.  

\begin{equation}
  {F}_{i} 
  = 
  \tbt{\cos\psi}   {\sin\psi}   {-\sin\psi}   {\cos\psi}  
  \label{eq:Fi}
\end{equation}


\end{itemize}


\end{document}
