\documentstyle{article}
%\input{html.sty}
\begin{document}
\newcommand{\aipspp}{AIPS++}
\title{The Generic Instrument: IV Specifications and Development Plan}
\author{T.J. Cornwell, NRAO}
\date{Version of 1995 May 9}
\maketitle

\def\MEGI{{\tt MEGI}}
\newcommand{\remark}[1]{{\small\begin{itemize}\em \item #1\end{itemize}}}

\newcommand{\ih}{\marginpar{{\em\small In hand}}}
\newcommand{\ok}{\marginpar{{\em\small OK in principle}}}
\newcommand{\nc}{\marginpar{{\em\small Not yet clear}}}
\tableofcontents

\section{Introduction}

This note follows on from previous memos by Cornwell (1995a, 1995b) on
the Hamaker, Bregman, and Sault (1995) calibration formalism. Here I
specify the capabilities demanded of the calibration and imaging
software and lay out a development plan to cover activities for the
next 12-18 months.  I have worked from memo 115 in the AIPS++ User
Specifications Memo series, authored by Bob Hjellming. My method has
been as follows:

\begin{enumerate}
\item Copy those relevant sections from that memo,
\item Remove items that are irrelevant or out-dated,
\item Consolidate items into re-statements relevant to the generic
instrument
\end{enumerate}

I have tried hard to adhere to the original specifications, but have
simplified where possible. As an example of simplification, I have tried
to remove most statements that VLBI processing is special in some way.
Similarly, I have assumed that mosaic processing will be standard.

Next I enumerate the priorities from NFRA (for WSRT), ATNF (for ATCA),
and NRAO (for the VLA and VLBA). All of these fit within the larger
set of specifications so the most important thing is the priority
given to these items.

Finally, I give a development plan designed to cover work in
the Synthesis area within AIPS++ over the next 12-18 months.

\section{Nature of Instrumental Data}

With only a few modifications, the Hjellming memo has the following
requirements on the nature of the instrumental data to be
addressed by AIPS++.

\begin{enumerate}
\item All data from the GI should be assumed to potentially (but not
necessarily) involve full measurement of the electromagnetic field
involving all four Stokes parameters
\ih
\item Multiple frequency bands may be simultaneously observed 
(e.g., for observing multiple lines simultaneously or multi-frequency
synthesis), with variable numbers of channels in each band
\ih
\item Single dish and synthesis array data will be handled on an
equal footing.
\ih
\item Rapid time switching of polarizations, frequencies, 
and pointing centers must be allowed.
\ok
\item Polarization measurements may be time switched if all
polarization measurements are not obtained simultaneously
\ok
\item Data combinations for different observations may have
different numbers of spectral channels and channel widths which may
need to be accommodated within single data sets
\ih
\item Observations should be regarded as mosaiced by default.
Mosaicing observations may have many ($\sim$1000) pointing centers
\ih
\item Single dish observations are special cases of mosaiced data.
\ok
\item Beam switched or multibeam systems on single dishes must be accomodated.
\ih
\item If available from the on-line data archive, instrumental 
performance and meteorological data must be associated with
instrumental data sets.
\remark{Little thought has been given to this requirement as yet.}
\ok
\item Meaningful error measures or estimates 
should be regarded as standard in an observation.
\remark{This will need substantial feedback to those originating data}
\ok
\item Antenna size, system temperatures, and frequency band-passes
may differ widely.
\ok
\item The integration time may vary from antenna pair to antenna pair.
\ok
\item VLB antennas in space will require support for orbital position
dependence including acceleration terms. An external emphemeris will
be required.
\ok
\end{enumerate}

\section{Data Correction and Calibration}
\subsection{General}
\begin{enumerate}
\item Data should be selectable in terms of identification with a 
particular type of calibration observation
\remark{This is unclear: we presume that it means that for example
phase calibration observations must be distinguished from
bandpass calibration observations. If so this is a far-reaching
problem, extending back into the observing system.}
\item Calibration should be made as generic as possible, with 
telescope-specific methods kept to a minimum
\ih
\item Both standard and user-defined models of data behavior should be
usable in determining calibration information from data sets
\ok
\item Instrumental behavior that affects calibration should be
integrable in the calibration process through a mixture of
parameterized functions and models in tabular form
\ok
\item Data correction based upon standard and user-defined functions,
with user supplied parameters, should be possible
\ok
\item Calibration and correction of data should be reversible, with
the capability to apply calibration/correction information either 
 ``on-the-fly'' during processing, or ``once and for all'', creating new,
calibrated data sets
\remark{The current design has an additional column in the
MeasurementSet for corrected coherences.}
\ok
\item Calibration/correction of data should be possible from derived
tables of instrumental parameters (e.g., system temperature vs. time,
gain vs. elevations), with derivation of such tables from calibration
observations or from on-line measurements
\ok
\item The calibration process should include flexible averaging of
calibration data and application with interpolations or weighted
averaging, all under control of the user
\ok
\item Cross-calibration from different instruments should be possible
(e.g. flux scale, pointing) particular when data from different
arrays are to be combined
\ok
\item Model fitting should be possible in both the image and u-v
planes, and it should be possible to use the resultant models for
further calibration and self-calibration
\remark{Model fitting in the u-v plane is not part of the 
\MEGI\ formalism. It must be performed outside the \MEGI\ framework.}
\item There must be simulation programs for single dish, interferometer, 
and mosaicing data bases for both planning and comparison of data with
models - with optional error generation for thermal noise, pointing
errors, primary beam errors, atmosphere, antennas surface errors,
beam-switching for total power, etc.
\ok
\end{enumerate}

\subsection{Interferometer Data}
\begin{enumerate}
\item Transfer of calibration matrices from one observation to another
should be possible and easy.
\ok
\item Redundancy in data (possibly including crossing points) should
be used whenever possible as an additional constraint on calibration
and self-calibration
\nc
\item Determination of, and application of corrections for, closure errors
should be possible with flexible averaging of input closure information
\ok
\item Fringe fitting for a range of spectral channels and fringe rates
should be possible by baseline, as well as globally by antenna
\ok
\item Spectra calculation from complex summing of visibilities 
in each spectral channel for user-specified positions in the field of view
\ok
\item Interferometric pointing, baseline, and beam pattern fitting
and related analysis
\remark{These lie somewhat out of the \MEGI\ framework and should be
regarded as operations on Jones matrices. For example, for antenna
position fitting, one would presumably use the \MEGI\ framework to
derive a G-matrix object, and then one would fit the phases in that
object to find antenna positions.}
\item Application and de-application of astrometric/geodetic
correction factors with complete and reversible histories
\ok
\item Calibration of data for effects of the ionosphere, utilizing
data at multiple frequencies and/or external data on variations of
electron content
\ok
\item Self-Calibration for non-isoplanicity must be possible
\remark{The \MEGI\ framework allows it but as far as we know, there
are no demonstrated algorithms}
\ok
\item Determination and correction for pointing errors, and errors in
beam shape, using mosaic self-calibration techniques, will be important
\ok
\item For spectral line sources one can do amplitude calibration
with auto-correlation spectra plus calibration at one antenna
\ok
\item Accurate Doppler correction for each spectral channel is essential
\ok
\item For polarization calibration, one must be able to determine
both source polarization structure and instrumental polarization
(``D-term'' self-calibration)
\ih
\item Full phase calibration is an iterative process involving 
limits set by: astrometry, geodesy, and weak source imaging/detection,
therefore one needs:
  \begin{enumerate}
  \item very accurate geometric models, typically to at least 1/10 of
  a wavelength accuracy
\remark{Possible eventually with the Measure system, currently being
designed and implemented by Wim Brouw. Models accurate at this level
will become available in late 1996.}
\ok
  \item knowledge of location of the Earth's pole and UT1, both of
  which are generally known only after astrometric/geodetic analysis
  \remark{Also possible with the Measure system. Available mid-1996}
\ok
  \item values of ionospheric delay as determined from measurements
  at simultaneous frequencies, or external measurements of ionospheric
  electron content
\ok
  \item measurement of properties of troposphere dry terms (from
  surface meteorological measurements) and wet terms (Kalman filtering,
  GPS multi-frequency satellite measurements, WVR)
\ok
  \item instrumental delays as determined from phase calibration
  signals
\ok
  \item knowledge of non-rigidity of the earth due to earth tides
  and atmospheric loading
\ok
  \end{enumerate}
\end{enumerate}

\subsection{Data Editing}
The requirements for data editing lie more in the domain of
visualization.
\begin{enumerate}
\item Data display and editing should be seen as generic tools
applicable to single dish, interferometer, and other forms of data
\ok
\item Data visualization for evaluation and editing purposes should
be seen as an integral, or closely coupled, aspect of the data system
\ok
\item It should be possible to do interactive editing based upon
display, with ``zoom'' or magnification, and menu selection of editing
options
\ok
\item Various ``viewing strategies'' should be available
  \begin{enumerate}
  \item For interferometer data, baseline by baseline display (with 
  magnification of local areas) and interactive editing (including 
  multiple, simultaneous baselines)
  using both Intensity-time-baseline displays and Intensity displays
  in u-v plane
  \item Displays of spectra and spectral cubes aggregated in various
  ways (spectra vs. time, averaging in time, averaging of channels)
  \item Selection of data by specifying windows in space and/or time
  \item Selection of arbitrary cuts through data (e.g. circular,
  radial, or a user-defined locus) through selected data coordinates
  \item Display of expanded data aggregates (e.g., pointing and
  clicking on an average multi-channel region of data to show the
  component spectrum
  \item Comparison displays of generic model data (from fitted
   components) with observed and/or processed data, including display
   of data with model subtracted or divided
  \end{enumerate} 
\ok
\item Data editing should be reversible, with the capability to store,
apply, and un-do editing information
\ok
\item Data editing should be possible on the basis of monitor/observing 
log data
\ok
\item Editing should be possible from ``consistency check''
information, particularly where there is redundancy or (for 
interferometric data) where there are crossing points in the u-v plane
\ok
\item It is desirable to have parameter-driven, automated 
flagging for large data sets
\ok
\item Editing must be possible
based upon difference between data and models generated
during self-calibration
\ok
\item Data editing based upon recognition of interference patterns
in intensity-time-frequency data is very important, particularly for
low frequency observations
\ok
\end{enumerate}

\section{Imaging and Image Processing}

In this section we consider the formation of images from edited,
calibrated data.  While this is mainly image computation and
deconvolution, it must be remembered, that for the user, imaging and
image deconvolution is an integral part of the process of data
inspection/editing, calibration, imaging, self-calibration, data/image
display, spectrum/time/image analysis, and production of hard copy for
publication purpose.  This process must be well integrated for the
convenience of the user.  It should be possible to easily
``mix-and-match'' self-calibration, data transformation, and
de-convolution ``tools'', for example, using CLEAN to deconvolve in
the early stages, and maximum entropy later on when CLEAN begins to be
less useful.  This is related to the need to make self-calibration use
a generic model, which could be a table of CLEAN-components, a table
of Gaussian components, or an image.

\subsection{Image and Spectral Image Formation}
\begin{enumerate}
\item Image construction using u-v data sets must be possible
with a range of capabilities
  \begin{enumerate}
  \item Computation of ``dirty'' images and point spread functions
  by 2-D FFT of selected data with user
  control of data selection, gridding algorithm and its parameters,
  and image parameters (image size, cell sizes, polarization)
\ih
  \item Flexible computation of data cubes where the third axis is
  frequency/velocity or time
\ok
  \item Simultaneous, multiple field imaging
\ok
  \item Direct Fourier transform imaging of arbitrary (and usually small)
  size fields 
\ih
  \item Imaging after subtraction for sources
\ok
  \item Imaging of spectral line data sets with continuum subtraction
  based upon continuum data, or continuum models
\ok
  \item Estimation and input of zero-spacing flux density and
  appropriate weighting
\ok
  \end{enumerate}
\item Mosaic image construction using mixture of u-v data sets and
single dish data for multiple antenna pointing centers
  \begin{enumerate}
  \item Linear combination of pre-deconvolved images, weighting
  determined by primary beam
\ok
  \item Linear mosaic algorithm with linear deconvolution (MOSLIN in SDE)
\ok
  \item Non-linear (MEM-based) mosaic algorithm (VTESS, UTESS in AIPS,
   mosaic in SDE)
\ok
  \item Cross-calibration (enforced consistency) between data taken
  with different instruments (flux scale, pointing)
\ok
  \item Pointing self-calibration to determine corrections for both
  single dish and visibility data
\ok
  \item Non-coplanar baselines mosaicing allowing for sky curvature
\ok
  \item Self-calibration and editing of all pointings in one processing
  step
\ok
  \item Capability to determine the primary beam(s) from a mosaic
  image and its related data sets
\ok
  \item Ability to deal with any primary beams in different forms
  (analytic 1- and 2-D, tabular), including user modification of primary
  beam models
\ok
  \end{enumerate}
\item Imaging using multiple-frequency data sets and a user-defined
model for spectral combination ``rules'' must be possible
\ok
\item Imaging computation should generally take multiple data sets
where this makes sense
\ok
\item Imaging data selection should flexibly allow use of data
sub-sets, with data selection based upon time, antenna, frequency, and
ranges of other data (including monitor data)
\ok
\item Non-coplanar baselines imaging (dragon in SDE)
\ok
\item Imaging wide fields large than the isoplanatic region
\ok
\item Near field imaging of nearby objects like comets and asteroids
\ok
\item Fringe-rate imaging
\ok
\end{enumerate}

\section{Priorities for development}

While the previous sections reflect the specifications for
AIPS++ functionality in the area of synthesis processing, the
priorities for development are not given. In this section,
I discuss these priorities.

\subsection{Completion of MEGI design}

Cornwell and Wieringa (1996) describe the design of the MEGI. The
following have yet to be completed:

\begin{enumerate}
\item Scalar version for {\em e.g.} RR alone or RR and LL only.
\item Full persistence of objects
\item Optimization for speed, including gridding and Fourier transformation
\item Cross-calibration
\end{enumerate}

\subsection{NFRA}

NFRA wants to use AIPS++ for data analysis for the new WSRT on-line
system TMS. TMS is expected to debut in August 1996. NFRA needs a
commitment from AIPS++ to support such use of AIPS++. As outlined
by Jan Noordam, the 
%\htmladdnormallink{
priorities
%} {http://www.aoc.nrao.edu/aips++/docs/specs/116.ps} 
are:

\begin{enumerate}
\item AIPS++ port to HP/UX
\item Fill to AIPS++ Synthesis MeasurementSet from WSRT data
format
\item Simple calibration, editing and imaging of WSRT
data, controllable from TMS
\item Polarized sky models including XX,YY,XY and YX as
well as the standard I,Q,U,V
\item Parametrized source components. Initially only for existing
calibrator models.
\item Subtraction of known sources from coherences
\item Simple data statistics: averages, rms, etc.
\item Visualization of coherence data
\end{enumerate}

\subsection{ATCA}

Mark Wieringa outlined 
% \htmladdnormallink{
priorities for ATCA calibration
%}{http://www.aoc.nrao.edu/aips++/docs/specs/117.ps}
:
\begin{enumerate}
\item A Solver for bandpass and gain using the parallel hand correlations.
\item A Solver for polarization leakage, gain and optionally 
source polarization using all four correlations.
\item Selfcal Solvers using either single Stokes (I) correlation data or 
multiple correlations.
\item A Corrector for each of bandpass, leakage and antenna/i.f.-gain and
a versioning scheme for either the Correctors or their underlying tables.
\end{enumerate}

There is no specific ``drop-dead'' date attached to these priorities
since Miriad currently can be used.

\subsection{BIMA}

Peter Teuben (1996) has outlined the special needs of BIMA.

\begin{enumerate} 
\item Support for time-sliced fashion polarization measurements, measuring circular 
polarization (LR and RL) with a quarter-wave plate. See also Wright (1995a) for a
discussion on some of the possibilities.
\item Deconvolution of mosaiced fields, including pointing corrections. Also 
adding single dish data to interferometry data. 
\item VLBI: BIMA regularly participates in mm VLBI experiments. The phased array data are 
currently processed offline using standard VLBI techniques (Mark-xxx, ref. xxx), and
whenever AIPS++ will provide VLBI data processing, BIMA should be able to use them without any major problems. 
\item Heterogenous array elements.
\item Unusual correllator modes: The correlator can be configured in many modes, and produces DSB 
data with a small (< 8) number of windows with different settings of the IF. An interesting method 
to calibrate DSB data is to use the generally much slower varying gain ratio (phase difference and 
amplitude ratio).
\item Offline phase corrections, using total power measurements (see Wright 1995b), examplify 
one of the many ways in which calibration needs to be flexible. 
\end{enumerate}

MIRIAD employs a very general visibility file format, where
correlations (both cross- and auto) are tagged with a rich set of
(name based) variables. Variables can be multi-dimensional of any of
the basic types (variables can change dimension in a dataset, in
principle even type).  Obviously, like in the FITS community, these
variables need to be registered and their meaning clarified. Currently
MIRIAD knows about 95 variables.  Although it would be ideal that the
telescope data be directly written in native AIPS++ format, for the
foreseeable future, a conversion program will be used.

\subsection{VLA}

A mail message from Michael Rupen dated Sept 19, 1995, gives the following
priorities for VLA software development.
\begin{enumerate}
\item Interpolating bandpass solutions.
\item Ionospheric corrections from GPS.
\item D-term self-calibration (with and without time-variability).
\item Mosaicing.
\item High dynamic range imaging (Briggs' NNLS algorithm)
\item Automated flagging.
\end{enumerate}

In general, NRAO will use whatever system is appropriate to acheive a
given functionality.  In some cases, AIPS is still the chosen route,
whereas in others AIPS++ is preferred. The first and second items are being
addressed within AIPS.

\subsection{VLBI}

Tony Beasley has summarized the special requirements for VLBI:

\begin{enumerate}
\item High precision quantities
\item Version typing to allow tracking of e.g. interferometer models used
\item Multi-file datasets (greater than 2 Gbyte and perhaps spread over many
disks)
\item Variable and possibly unequal integration times
\item Provision for tied arrays
\item Diversity of antenna mounts
\item ``In the beam'' phase referencing
\end{enumerate}

In addition, tasks for the following will be needed:

\begin{enumerate}
\item Data readers for various formats
\item Sophisticated model for correlator effects e.g. decorrelation, state-count
corrections.
\item Absolute amplitude calibration is more important than in connected
element interferometry. There are various approachs, all of which must be
supported.
\item Fringe fitting, both antenna and baseline-based.
\item External data calibration
\item Phase-cal information
\item Polarization calibration, including all known calibration algorithms
\item Velocity correction for fringe-rotation
\item Pulsar binning and gating
\item Source fitting
\end{enumerate}

\section{Development plan}

Here we cover the parts of the development plan appropriate to
synthesis imaging.  The goals of this development plan are to address
pressing needs of the consortium partners and to push forward into new
application areas. It is expected that this development plan will be
revised from time to time to incorporate, for example, VLBI processing
when appropriate.

\begin{tabbing}
++++++++++++++++++++++++++++\=++++++++++\= \kill
AIPS++ port to g++     \> June 96 \> NFRA \\
Final version of WSRT Filler \> July 96 \> Olnon \\
Spectral support       \> July 96 \> Wieringa \\
Cross calibration and editing \> July 96 \> Cornwell \\
ComponentSkyModel \> July 96 \> Marson \\
TMS commissioning      \> Aug 96 \> de Vos \\
VLA 90cm imaging       \> Oct 96 \> Cornwell \\
Mosaicing              \> Dec 96 \> Cornwell
\end{tabbing}

Some notes:
\begin{description}
\item[Mosaicing] support is a way to test the framework for 
inclusion of image plane effects.
\item[Spectral support] means that the \MEGI\ framework must be
extended to allow multiple spectral channels. This is straightforward,
\item[VLA 90cm imaging] will be accomplished using a program similar
in principle to dragon in SDE.
\end{description}

\section*{References}

\noindent 
%\htmladdnormallink{
Briggs, D.S., 1995, Ph.D. thesis, New Mexico Institute of Mining and Technology
%}{http://www.aoc.nrao.edu/ftp/dissertations/dbriggs/diss.html}

\noindent Cornwell, T.J., 1995, AIPS++ Implementation Note 183.

\noindent Cornwell, T.J., 1995, AIPS++ Implementation Note 184.

\noindent Cornwell, T.J., and Wieringa, M.H., 1996, AIPS++ Implementation Note 189.

\noindent Hamaker, J.P., Bregman, J.D., and Sault, R.J., 1995,
{\em Understanding radio polarimetry: I Mathematical foundations},
submitted to A\&A.

\noindent 
%\htmladdnormallink{
Teuben, P, 1996, BIMA calibration requirements for AIPS++
%}{http://www.astro.umd.edu/~teuben/aips++/bima.html}
\end{document}