\documentstyle[10pt]{article}
\begin{document}

\newcommand{\aipspp}{{\sc aips++}}

\title{Note 211: Alternative window managers for Green Bank}
\author{J.P. McMullin \\ {\em jmcmulli@nrao.edu}}
\date{Sept 22, 1997}
\maketitle

\section{Movtivation}

CDE (Common Desktop Environment) is the native window manager for NRAO, Green
Bank. This window manager tends to require large amounts of memory and to
consume much of the available color allocation. This note explores
alternative window managers which will provide the best environment for 
running AIPS++. 

\section{The Players}

Four different window managers were considered:

CDE: Common Desktop Environment - a sophisticated customizable desktop 
environment.

FVWM2: F\footnote{choose your favorite f word} Virtual Window Manager - a 
sophisticated customizable desktop environment. The '2' designates the current 
version in release.

TWM: Tom's Window Manager - one of the early window managers and consequently
one of the more primitive. Does not support desktops.

OLWM: Open-look Window Manager - SUN's old window manager. Does not support
desktops.

\section{Memory Breakdown}

The following table is a breakdown of the memory use of the different 
managers. The column outputs are the amounts of used swap space (in
kilobytes) for different amounts of activity. '2 xterms' designates
the memory usage for having two available xterms (in the case of olwm
and fvwm2 this also includes a console terminal for network and status
messages) while the subsequent rows indicate the addition of more 
xterminals, a netscape session, and an AIPS++ session.

\begin{tabular}{lcccc}
	 &CDE 	 & fvwm2 	& twm 	& olwm \\
	 &	 &		& 	& 	\\
2 xterms & 28352 & 24680        & 24464 & 23512 \\
+2 xterms& 30176 & 25832	& - 	& - \\
+netscape& 34480 & 30032	& -	& - \\
+aips++  & 50560 & 46384	& - 	& - \\
\end{tabular}

I only compared CDE and fvwm2 in all of the categories since they were
the only two which supported desktops, considered an essential
feature. Overall, twm and olwm do not represent great memory savings for
significantly less performance.

In addition, the 'top' utility was also used to assess the memory usage.
CDE requires a large amount of resident memory to pay for its sophistication
while FVWM2 offers a similar package with a leaner presence.

\begin{verbatim}
  PID USERNAME PRI NICE  SIZE   RES STATE   TIME   WCPU    CPU COMMAND

 8880 jmcmulli  33    0 2112K 1704K sleep   0:00  0.03%  0.04% fvwm2
 8880 jmcmulli  34    0 2672K 2152K sleep   0:01  0.05%  0.05% twm
 9660 jmcmulli  34    0 2208K 1920K sleep   0:00  0.49%  0.38% olwm
 9817 jmcmulli  34    0 6376K 5528K sleep   0:01  1.01%  0.24% dtwm
\end{verbatim}

Beyond memory, the color allocation within CDE has also caused problems
when interacting with programs within AIPS++ (e.g. aipsview). This is
true even without other color intensive processes (e.g. netscape) present.

Overall, the memory use and color consumption of CDE, combined with the
disparity between the canonical AIPS++ machine and the typical low-memory
IPC's generally in use at Green Bank strongly suggests that the CDE is
actually the {\it worst} window manager system for those requiring high
performance in their AIPS++ applications.

\section{Switching from CDE}

Switching from CDE to an alternative window manager requires three basic
steps: 1) the addition of two files to the home directory; this is true for 
all of the alternative window managers discussed here, 2) the addition of
a directory into your PATH and LD\_LIBRARY\_PATH, and 3) breaking out
of CDE.

\subsection{Adding the required files}

The two files required are a .xinitrc file and a .wmrc file (window manager
rc file, eg. .fvwm2rc, twmrc, etc.).

The .xinitrc file dictates the starting windows and window manager for your
X session. The following is an example file:

\begin{verbatim}
#!/usr/bin/tcsh
#
# loads local X resource information
xrdb -load $HOME/.Xdefaults 
# Console window for important network and status messages
xterm -C -title console -name console -geometry 98x6-0-0 -sb & 
# starts up 2 xterminals and puts them in the upper right and
# lower left corners of the screen
xterm -geometry -0+0 -name xterm -sb &
xterm -geometry +0-0 -name xterm -sb &
# set up a clock in the lower right
clock -geometry 100x100-30-110 &
sleep 4
# setup a nice background image if you have one
# don't get too fancy or it will consume all of the colors!!
xloadimage -zoom 110 -onroot $HOME/img/bluepearl.gif & 
# This is the window manager that I want to use.
# Alternatively, I could have specified twm or olwm.
fvwm2
\end{verbatim}

The second file required is the .wmrc file (e.g. .twmrc, .fvwm2rc, etc.). 
The .fvwm2rc file sets the configuration of desktops, the attributes
of various processes, etc.
fvwm2 appears to offer the best overall system in that it is relatively svelte
in its memory and color usage, yet allows a great deal of freedom in
setting up a personal work environment. In addition, fvwm2 serves as a
platform for other managers, if desired; the left mouse button, selecting
on Exit FVWM, offers a startup for twm and olwm.
As a result, I offer only a .fvwm2r\_sample file for public use. 

The .fvwm2rc\_sample file may be copied from $\sim$jmcmulli to your home 
directory.  Information on the various parameters can be found locally (the 
fvwm2 manual) at http:www.gb.nrao.edu/$\sim$jmcmulli/fvwm.man.html; the FVWM 
home page is located at: http://www3.hmc.edu/$\sim$tkelly/docs/proj/fvwm.html.

\subsection{Adding to your PATH and LD\_LIBRARY\_PATH.}

Two directories of standard X-resources will be needed to run FVWM2: 
/opt/local/X11R5/bin and /opt/local/X11R5/lib.

If you use the csh,tcsh, add the following lines to the bottom of your
.cshrc, .tcshrc file.

\begin{verbatim}
setenv LD_LIBRARY_PATH "$LD_LIBRARY_PATH":/opt/local/X11R5/lib
set path=($path /opt/local/X11R5/bin )
\end{verbatim}

If you use the bash, add the following lines to the bottom of your
.bashrc file.

\begin{verbatim}
export LD_LIBRARY_PATH=($LD_LIBRARY_PATH /opt/local/X11R5/lib)
export PATH=($PATH /opt/local/X11R5/bin)
\end{verbatim}

\subsection{Getting out of CDE and into FVWM2}

After the previous setups are finished. Logout of your CDE session. Then,
use the mouse to show the Options menu on the CDE login screen. Select
Command Line Login.

Hit $<$Return$>$ to get to the login prompt. Type in your userid and password.

Once you're logged in, begin your new session by typing 'startx'.

You're in. Have fun.

If you are interested in seeing fvwm2 in action before committing to the
change, please stop by my office and investigate.

\end{document}
