\documentstyle{article}
\setlength{\textwidth}{6.5in}
\setlength{\evensidemargin}{0.0in}
\setlength{\oddsidemargin}{0.0in}
\parskip=5truept
\begin{document}
\newcommand{\aipspp}{AIPS++\ }
\newcommand{\defline}[1]{\cline{1-5} \multicolumn{5}{|l|}{#1} \\ \cline{1-5}}

\newcommand{\definetable}[2]
{
        %\clearpage
	\vfill
        %\eject
	%\subsection{#1}
        \vspace{0.15in}
        \small

	\begin{tabular}{|l|p{1.25in}|l|p{.9in}|p{1.4in}|}
	\hline
	\multicolumn{5}{|c|}{\bf #1}\\ 
	\cline{1-5} 
        \multicolumn{1}{|c|}{Name}&\multicolumn{1}{|c|}{Format}&
        \multicolumn{1}{|c|}{Units}&\multicolumn{1}{|c|}{Measure}&
        \multicolumn{1}{|c|}{Comments}\\ \cline{1-5}
        %#2
        \hline
	\end{tabular}
}

\title{Definition of MeasurementSet \break \aipspp\ Note 191}
\author{M.H. Wieringa, ATNF\\T.J. Cornwell, NRAO}
\date{October 4, 1996}
\maketitle

\tableofcontents
\section{Summary}
A MeasurementSet is an \aipspp\ Table containing data from a telescope.
Telescopes all record data in their own way, the MeasurementSet layout
describes how these measurements are to be stored within the \aipspp\ system.
This document describes the predefined columns and keywords (including
subtables) of the MeasurementSet (MS). The predefined items include
required and optional ones.
In addition to the predefined columns, MeasurementSets can contain
instrument specific columns, keywords or subtables. These additional
items would only be accessed from instrument specific code.

\section{Details}

\begin{itemize}
\item The purpose of the columns and tables described here is to store
information that one can expect to arrive from a telescope.
Calibrated data will be the DATA column in a calibrated MS that is
otherwise simply a reference to the original MS (note that the UNITs
of the DATA column may therefore be different in original and
calibrated MSes). Predicted data will similarly show up as a DATA
column in an MS.

\item There is only one subtable of each type and
that is stored as a keyword of the table. This simplifies selection since
it can be done on a table (e.g. SOURCE NAME is '3C84') and the keys then
used to select from the main table.

\item We want to keep as many columns as possible common to both
Single Dish and Synthesis. Furthermore, we regard most of the columns
identified here as required. It simplifies code tremendously and the
Miriad storage manager avoids significant disk space usage. However,
we have identified a couple of cases where following these guidelines
would result in too much wasted storage and introduced some optional
columns to cope with this (e.g., FLOAT\_DATA for single dish).

\item The model is that each ANTENNA has an unrestricted number of FEEDs.
For each FEED there are some RECEPTORS, probably one per polarization state
(e.g. R and L or X and Y). A number of SPECTRAL\_WINDOW setups are allowed.
Different setups are expected in different rows of the Main table.
Thus, for example, different VLA IFs show up as different rows.

\item We think that PHASED\_ANTENNAS and PHASED\_FEEDS table will be needed
but since these are quite specialized, we have refrained from specifying
contents. We think that these should be implemented first as TELESCOPE specific
tables (e.g. NRAO\_GBT\_FEEDARRAY) and then a common format agreed upon.
Such tables should probably index into the ANTENNA and FEED tables to get
information.

\item The SOURCE table contains information that is often not present
in on-line formats, therefore we have made the use of the table
optional. It could be added during later processing to organize the 
data by source. The FIELD table is the place to describe pointings.

\item The CORRELATOR table is still unspecified. Instruments like the
VLBA will require it, but at present it will usually be empty.

\item The PULSAR table appears to be very telescope specific so we haven't
described any format. Perhaps this possibility would provide a good impetus
for the pulsar community to make common cause. 

\item Non-standard information may be added in various ways as per Bob
Garwood's proposal:
\begin{itemize}
\item As a new column in an existing table e.g. NS\_NRAO\_GBT\_WHATEVER
\item As an e.g. NRAO-approved column NRAO\_GBT\_WHATEVER or NRAO\_WHATEVER
\item As a Project approved column WHATEVER.
\end{itemize}

\item Units and Measures. All columns representing physical quantities
should be specified in the correct units, as specified in the table
descriptions. Columns representing \aipspp Measures should have a
MEASURE keyword giving the \aipspp Measure name minus the initial 'M'
(e.g., DIRECTION or EPOCH) and a MEASURE\_REFERENCE keyword giving the 
Measure enum type as a String (e.g., "J2000" or "UTC")

\end{itemize}


\section{MeasurementSet classes}
\par
The MeasurementSet is coded as a series of related classes. Each
subtable follows the same basic scheme as the MAIN table. 
The namespace in each table is separate. For each
(sub)table there is a separate class which provides various conversion
and inquiry functions for the columns and keywords and defines the
required set of columns and keywords. The main MeasumentSet class
provides access to the subtables via these classes.  E.g.,
MeasurementSet has a member function antenna() which gives access to
the ANTENNA subtable. More detailed documentation on these classes can
be found in the on-line documentation 
(directory: code/trial/implement/MeasurementSet).

The definite descriptions of the MS tables are contained within the
\aipspp class documentation files in the MeasurementSet
module. For each of the MeasurementSet
tables there is a file (e.g. MSMainEnums.html, MSAntennaEnums.html) giving
a description for the enum name (e.g., DATA, POSITION) used for a
particular column or keyword.  
Examples on how to use these classes to write a filler for a
particular instrument can be found by looking at existing fillers,
e.g., uvfitsfiller.

\clearpage
\section{MS layout}

There is a MAIN table containing a number of data columns and keys
into various subtables. There is at most one of each subtable. The
subtables are stored as keywords of the MS.

\begin{tabular}{|l|p{1.5in}|p{1.6in}|} \hline
\multicolumn{3}{|c|}{\bf Subtables}\\ \cline{1-3}
\multicolumn{1}{|c|}{Table}&\multicolumn{1}{|c|}{Contents}&\multicolumn{1}{|c|}{Keys}\\ \cline{1-3}
ANTENNA& Antenna characteristics & ANTENNA\_ID, ARRAY\_ID \\
ARRAY  &         Array characteristics & ARRAY\_ID \\
CORRELATOR & Correlator setup & CORRELATOR\_ID\\
FEED  &          Feed characteristics & FEED\_ID, ANTENNA\_ID, ARRAY\_ID, TIME, SPECTRAL\_WINDOW\_ID \\
FIELD  &         Position etc for each pointing. & FIELD\_ID \\
FLAGGING\_LOG  &  Log of flagging operations & ANTENNA\_ID, ARRAY\_ID, FEED\_ID, SPECTRAL\_WINDOW\_ID, FIELD\_ID, TIME, INTERVAL \\
OBSERVATION  &Observer, Schedule, etc & OBSERVATION\_ID \\ 
OBS\_LOG  &Log from on-line system & OBSERVATION\_ID, TIME \\
SOURCE  &        Positions, etc for each source & SOURCE\_ID, SPECTRAL\_WINDOW\_ID, TIME \\
SPECTRAL\_WINDOW  &     Spectral window setups & SPECTRAL\_WINDOW\_ID \\
SYSCAL &         System calibration characteristics & FEED\_ID, ANTENNA\_ID, ARRAY\_ID, TIME, SPECTRAL\_WINDOW\_ID \\
WEATHER  &       Weather info for each antenna & ANTENNA\_ID, ARRAY\_ID, TIME \\
\hline
\end{tabular}

\par
Note that there are two types of subtables. For the first, simpler
type, the key (ID) is the row number in the subtable. 
Examples are ARRAY, FIELD, SPECTRAL\_WINDOW, OBSERVATION
and CORRELATOR. For the second, the key is a collection of parameters,
usually including TIME. Examples are ANTENNA, FEED, SOURCE, SYSCAL, WEATHER.
We think the Calabretta interpolation test is a good one: CORRELATOR
setups cannot be sensibly interpolated and thus a different ID
is required for each setup. For the interpolable tables, one needs
to decide what value is actually to be used. We think this belongs in
access routines independent of the MS.

\par
Notes: 
\begin{itemize}
\item All ID columns are zero-based, a value of -1 indicates that
there is no corresponding subtable present. 
\item All required columns
should be filled with suitable defaults if not actually used. 
\item For time dependent tables, a value of DBL\_MAX (defined by including
aips/Mathematics/Constants.h) for interval can be used to
specify non time dependent entries.
\item The layout of the CORRELATOR and FLAGGING\_LOG table is currently
unspecified. We need to gain some experience with specific
implementations of these before specifying a general format.
\end{itemize}

\definetable{MAIN table: Coordinates, Data, pointers and Flags}
{
\defline{\bf Columns}
\defline{\em Coordinate information}
ANTENNA1  &      Int    &          &              & First antenna \\
ANTENNA2  &      Int    &          &              & Second antenna \\
EXPOSURE  &      Double &       s  &              & The effective integration time \\
FEED1     &      Int    &          &              & Feed on ANTENNA1 \\
FEED2  &      Int       &          &              & Feed on ANTENNA2 \\
INTERVAL  &      Double &      s   &      & Sampling interval \\
PULSAR\_BIN &       Int &          &              & Pulsar bin number \\
SCAN\_NUMBER  &   Int   &          &        & Scan Number  \\ 
TIME  &         Double  &      s   &     EPOCH    & Midpoint of the integration. \\
TIME\_EXTRA\_PREC & Double & s       &     & add to TIME for
extra precision (optional) \\
UVW       &   Double(3) &      m   &         UVW  & UVW coordinates \\
\defline{\em Pointers into subtables}
ARRAY\_ID  &      Int   &          &        & \\
CORRELATOR\_ID  & Int   &          &        & \\
FIELD\_ID  &      Int   &          &        & \\
OBSERVATION\_ID  & Int   &          &        & \\
PULSAR\_ID  &     Int   &          &        & \\  
SPECTRAL\_WINDOW\_ID  &  Int   &          &        & \\
\defline{\em Data}
DATA    &    Complex(NUM\_CORR, NUM\_CHAN) &  & & Complex visibility
matrix (synthesis arrays, optional)\\
FLOAT\_DATA & Float(NUM\_CORR, NUM\_CHAN) & & & Float data matrix
(single dish, optional) \\
SIGMA   &    Float(NUM\_CORR) &  & & Estimated rms noise for single channel \\ 
SIGMA\_SPECTRUM   &    Float(NUM\_CORR, NUM\_CHAN) &  & & Estimated rms noise (optional) \\ 
WEIGHT   &    Float &  & & Weight for whole data matrix \\ 
WEIGHT\_SPECTRUM   &    Float(NUM\_CHAN) &  & & Weight for each channel (optional)\\ 
\defline{\em Flag information}
FLAG    &    Bool(NUM\_CORR, NUM\_CHAN) &  &   &  The data flags \\ 
FLAG\_HISTORY & Bool (NUM\_CORR, NUM\_CHAN, NUM\_HIS) & & & History of
flags (optional)\\
FLAG\_ROW &    Bool  &  &   &  The row flag \\ 
}
\begin{description}
\item[Notes:]
\item[DATA, FLOAT\_DATA] At least one of these columns should be
present in any MeasurementSet. In special cases both could be present
(e.g., single dish data used in synthesis imaging or a mix of auto and
crosscorrelations on a multi-feed single dish)
\item[ANTENNA1, ANTENNA2] For Single dish ANTENNA1=ANTENNA2
\item[PULSAR\_BIN] For a pulsar the correlations are assumed to be measured for a
limited number of pulse phase bins. This is the particular bin for which this data was measured. 
\item[FLAG] An array of bools with the same shape as DATA. 
Data is flagged bad if FLAG is True.
\item[FLAG\_HISTORY] A set of NUM\_HIS flags for each data point, each
for a different category of flagging (e.g. on-line, calibration,
interactive, etc.)
\end{description}

\par Next we describe the various subtables. The columns are
categorized into Keys, Data description, Data, and Flags.

\definetable{ANTENNA: antenna characteristics}
{
\defline{\bf Columns}
\defline{\em Coordinate information}
ANTENNA\_ID &     Int     &          &              & Key \\
ARRAY\_ID &     Int     &          &              & Key \\
\defline{\em Pointers}
ORBIT\_ID    &    Int & & &  Orbit parameters \\
PHASED\_ARRAY\_ID & Int   &  & &   Phased array \\
\defline{\em Data}
DISH\_DIAMETER &  Double &  m  &      & Diameter of dish\\
MOUNT  & String &  & &   Mount type:alt-az, equatorial, X-Y, orbiting, bizarre \\
NAME &       String &  & &  Antenna name \\
OFFSET & Double(3)  &   m &     POSITION & Axes offset of mount to FEED REFERENCE point \\
POSITION &   Double(3) &  m   &     POSITION &   Antenna X,Y,Z phase reference positions \\
%POSITION\_RATE &  Double(3) & m/s &   VELOCITY(?) & Rate of change of position. \\
STATION &        String & & &  Station name \\
}
\begin{description}
\item[Notes:]
\item[POSITION]In a right-handed frame, X towards the intersection of
the equator and the Greenwich meridian, Z towards the pole. The exact
frame should be specified in the MEASURE\_REFERENCE keyword (ITRF or
WGS84). The reference point is the point on the az or ha axis closest
to the el or dec axis.
\end{description}

\definetable{ARRAY: Array characteristics}{
\defline{Data}
NAME      &     String  &          &              & 
Name of array (e.g. 'VLAA', 'EVN', 'VLBA-PT'). Informational only \\
POSITION  & Double(3)   & m & POSITION & Array reference position
(optional), same frame as antenna positions. \\
}

\definetable{FEED: Feed characteristics}{
\defline{\bf Columns}
\defline{\em Keys}
ANTENNA\_ID &     Int &    &      & Antenna id\\
ARRAY\_ID  &      Int   &          &        & Array id\\
FEED\_ID &        Int &    &      & Feed id\\
INTERVAL &       Double  &      s     &       & Interval\\
SPECTRAL\_WINDOW\_ID &   Int &    &      & Spectral Window id\\
TIME &   Double  &      s  & EPOCH & Midpoint of validity range of parameters\\
\defline{\em Data description}
NUM\_RECEPTORS &  Int &    &      & \# receptors on this feed \\
\defline{\em Data}
BEAM\_ID &        Int    &    &      & Beam model\\
BEAM\_OFFSET &    Double(2, NUM\_RECEPTORS)    &     rad & DIRECTION & Beam position offset (on sky but in antenna reference frame).\\
PHASED\_FEED\_ID & Int   &    &      & Phased feed\\
POLARIZATION\_TYPE &       String (NUM\_RECEPTORS) &    &      & Type of polarization to which a given RECEPTOR responds. Probably
        R, L or X, Y.\\
POL\_RESPONSE &  Complex (NUM\_RECEPTORS, NUM\_RECEPTORS) &    &      & D-matrix i.e. leakage between two receptors\\
POSITION &    Double(3)  & m  &         POSITION & Position of feed relative to feed reference position for this antenna\\
RECEPTOR\_ANGLE &    Double (NUM\_RECEPTORS) &    rad &    & The reference angle for polarization. Converts into
        Parallactic angle in the Sky domain.\\
}
\begin{description}
\item[Notes:]
\item[FEED] A FEED is e.g. a single horn.  A FEED can have one or two 
RECEPTORs that respond to different polarization states. Feeds are
numbered from 0 on each separate antenna and for each
SPECTRAL\_WINDOW\_ID. So FEED\_ID should be non-zero only in the case
of feed-arrays, i.e., multiple simultaneous beams on the sky at the
same frequency and polarization. 
\item[SPECTRAL\_WINDOW\_ID] A value of -1 indicates the row is
valid for all spectral windows.
\item[BEAM\_ID] Points to an optional BEAM subtable with parameters (or
image) for the primary beam for this antenna with this
feed. No format specified yet for the BEAM table.
\end{description}

\definetable{FIELD: Field positions for each source}{
\defline{\bf Columns}
\defline{\em Keys}
FIELD\_ID &       Int &    &      & Field Id\\
\defline{\em Data}  
CODE &       String &    &      & Special characteristics of field \\
DELAY\_DIR    &      Double(2)  &      rad   &     DIRECTION & Direction of delay center. \\
DELAY\_DIR\_RATE     &     Double(2)  & rad/s &   & Rate of change of direction of delay direction \\
NAME &       String &    &      & Name of field \\
PHASE\_DIR    &      Double(2)  &      rad   &     DIRECTION & Phase center. \\
PHASE\_DIR\_RATE     &     Double(2)      &  rad/s   &     &    Rate of change of direction of phase center.\\
POINTING\_DIR &      Double(2)  &      rad   &     DIRECTION & Pointing center \\
POINTING\_DIR\_RATE   &    Double(2)    &    rad/s   &     &   Rate of change of direction of pointing center.\\
REFERENCE\_DIR &      Double(2)  &      rad   &     DIRECTION & Reference center \\
REFERENCE\_DIR\_RATE   &    Double(2)    &    rad/s   &     &   Rate of change of direction of reference center.\\
SOURCE\_ID &      Int    &    &      & Index in Source table\\
TIME &   Double  &      s  & EPOCH & Time origin for the directions and rates\\
}
\begin{description}
\item[Notes:]
\item[SOURCE\_ID] Points to an entry in the optional SOURCE subtable, a
value of $-1$ indicates there is no corresponding source defined.
\end{description}

%\definetable{FLAGGING\_LOG: flag information}{
%\defline{\bf Keywords}
%\defline{\bf Columns}
%\defline{\em Keys}
%\defline{\em Data description}
%\defline{\em Data}  
%\defline{\em Flags}
%}

\definetable{OBSERVATION: Observation information}{
\defline{\bf Columns}
\defline{\em Data}  
CORR\_SCHEDULE &      String(*) &    &      & Correlator script\\
OBSERVER &       String &    &      & Name of observer(s)\\
OBS\_SCHEDULE &       String(*) &    &      & Project Schedule\\
PROJECT &               String &    &      & Project identification string.\\
\defline{\em Flags}
}

\begin{description}
\item[Notes:]
\item[OBSERVATION] This contains information about the observation process.
In the first instance, this would be a record only. Eventually we may 
want to be able to interpret Schedules for processing
hints e.g. CALIBRATION\_GROUPs. The OBSERVATION\_ID is the row number.
\end{description}

\definetable{OBS\_LOG: Observation log information}{
\defline{\bf Columns}
\defline{\em Keys}
OBSERVATION\_ID  & Int   &          &        & Points to OBSERVATION table\\
TIME & Double & s &              EPOCH  & Timestamp for message\\
\defline{\em Data}  
MESSAGE &String &    &      & log message\\
}

\definetable{SOURCE: Source information}{
\defline{\bf Columns}
\defline{\em Keys}
INTERVAL &       Double & s &               & Interval\\
SOURCE\_ID &      Int &    &      & Source id\\
SPECTRAL\_WINDOW\_ID &   Int &    &      & Spectral Window id\\
TIME &   Double & s  &   EPOCH  & Midpoint of time for which this set of parameters is accurate\\
\defline{\em Data}  
CALIBRATION\_GROUP &      Int &    &      & \# grouping for calibration purpose\\
CODE &       String &    &      & Special characteristics of source, e.g. Bandpass calibrator\\
DIRECTION &       Double(2) &  rad  &    DIRECTION &        Direction (e.g. RA, DEC) \\
NAME &       String &    &      & Name of source as given during observations\\
POSITION  & Double(3) &   m  & POSITION & Position (e.g. for solar system objects) \\
PROPER\_MOTION &  Double(2) &  rad/s & &  Proper motion \\
SYSVEL   &       Double  &  m/s  & RADIAL VELOCITY & Systemic velocity at reference \\
}

\definetable{SPECTRAL\_WINDOW: Spectral window setups}{
\defline{\bf Columns}
\defline{\em Data description columns}
CORR\_TYPE & Int(NUM\_CORR) & & & Polarization of correlation \\
CORR\_PRODUCT &  Int(2, NUM\_CORR) &  & & (see below) \\  
IF\_CONV\_CHAIN & Int & & & The IF conversion chain \\
NUM\_CHAN   &     Int  &           &        & \# spectral channels \\
NUM\_CORR   &     Int  &           &        & \# correlations\\
\defline{\em Data}  
CHAN\_FREQ  &   Double(NUM\_CHAN)    &           Hz    &    FREQUENCY& Center frequencies for each channel in the data matrix.\\
MOLECULE    &     String  &  &  & Molecule observed (optional) \\
REF\_FREQUENCY     &   Double       &               Hz  &   FREQUENCY&        The reference frequency.\\
RESOLUTION   &     Double(NUM\_CHAN)         &      Hz      &  &
The effective noise bandwidth of spectral channels\\
TOTAL\_BANDWIDTH & Double & Hz &  & total bandwidth for this window \\
TRANSITION  &     String &     &     & Transition of molecule (optional) \\
REST\_FREQUENCY    &    Double  &    Hz  & FREQUENCY &  The rest frequency (spectral line) \\
}
\begin{description}
\item[Notes:]
\item[CHAN\_FREQ] Can be non-linear to allow for e.g. AOS
\item[CORR\_PRODUCT] Pair of integers for each correlation product,
specifying the receptors from which the signal originated.
To get the meaning of the values (0 or 1), see the POLARIZATION\_TYPE
in the FEED table. An example would
be (0,0), (0,1), (1,0), (1,1) for all correlations between whatever
the receptors on FEED1 and FEED2 are measuring
\item[CORR\_TYPE] An integer for each correlation product indicating
the Stokes type as defined in the Stokes class enumeration.
\item[IF\_CONV\_CHAIN] Identification of the electronic signal
path for the case of multiple (simultaneous) IFs. (e.g. VLA: AC=0,
BD=1, ATCA: Freq1=0, Freq2=1)
\item[NUM\_CORR] RR would be 1, RR, LL would be 2,
XX,YY,XY,YX would be 4, etc.
\item[RESOLUTION] The effective noise bandwidth of spectral channels
        (the frequency resolution of each channel), this
        may differ from the channel spacing (difference
        in frequency between the centers of adjacent channels).
        The Vector nature allows for variable-width channels.
\end{description}

\definetable{SYSCAL: System calibration}{
\defline{\bf Columns}
\defline{\em Keys}
ANTENNA\_ID &     Int &    &      & Antenna id\\
ARRAY\_ID  &      Int   &          &        & Array id\\
FEED\_ID &        Int &    &      & Feed id\\
INTERVAL &       Double  &      s     &   & Interval\\
SPECTRAL\_WINDOW\_ID &   Int &    &      & Spectral window id\\
TIME      &   Double  &      s  & EPOCH & Midpoint of time for which this set of parameters is accurate\\
\defline{\em Data description}
NUM\_RECEPTORS &  Int &    &      & \# receptors on this feed (1 or 2)\\
\defline{\em Data}  
PHASE\_DIFF &  Float & rad &    & Phase difference between receptor 2 and receptor 1\\
TCAL &       Float (NUM\_RECEPTORS)  &      K &  & Calibration temp\\
TRX  &       Float (NUM\_RECEPTORS)  &      K &  & Receiver temperature\\
TSYS &       Float (NUM\_RECEPTORS)  &      K &  & System temp\\
TCAL\_SPECTRUM &       Float (NUM\_RECEPTORS, NUM\_CHAN)  &      K &  & Calibration temp (optional)\\
TRX\_SPECTRUM  &       Float (NUM\_RECEPTORS, NUM\_CHAN)  &      K &  & Receiver temperature (optional)\\
TSYS\_SPECTRUM &       Float (NUM\_RECEPTORS, NUM\_CHAN)  &      K &  & System temp (optional)\\
\defline{\em Flags}
PHASE\_DIFF\_FLAG &   Bool  &    &      & Flag for PHASE\_DIFF\\
TCAL\_FLAG &  Bool &    &      & Flag for TCAL\\
TRX\_FLAG &   Bool  &    &      & Flag for TRX\\
TSYS\_FLAG &  Bool &    &      & Flag for TSYS\\
}


\definetable{WEATHER: weather station information}{
\defline{\bf Columns}
\defline{\em Keys}
ANTENNA\_ID &        Int &    &      & Antenna number\\
ARRAY\_ID  &      Int   &          &        & Array id\\
INTERVAL &       Double &        s  &      & Interval over which data is relevant\\
TIME &  Double  &   s  & EPOCH  & \\
\defline{\em Data}  
H2O &        Float    &     $m^{-2}$  &      & Average column density of water\\
IONOS\_ELECTRON & Float & $m^{-2}$    &      & Average column density of electrons\\
PRESSURE &       Float &  Pa  &      & Ambient atmospheric pressure\\
REL\_HUMIDITY &   Float &    &      & Ambient relative humidity\\
TEMPERATURE &    Float &  K  &     & Ambient Air Temperature for an antenna\\
WIND\_DIRECTION & Float  &       rad &      & Average wind direction\\
WIND\_SPEED &  Float &  m/s  &    & Average wind speed \\
}

\end{document}


