\documentstyle{article}
\begin{document}
\newcommand{\aipspp}{AIPS++}

\title{The Generic Instrument: V Design of Cross-calibration}

\author{T.J. Cornwell, NRAO\\M.H. Wieringa, ATNF}
\date{1996 May 6}
\maketitle

\tableofcontents

\section{Introduction}

A previous memo described calibration and imaging for the generic
Interferometer (Cornwell and Wieringa, 1996) using the measurement
equation developed by Hamaker, Bregman and Sault (1995) (see also
Noordam, 1995). Most attention has been paid to self-calibration
whereby the derived calibration information need only be applied to
the source from which it was derived. The more general case is to
derive calibration on one source or set of sources, and then apply it
to another source.  The purpose of this memo is to sketch the form of
the design for this form of calibration, that we will call {\em
cross-calibration}. This is a sketch only and, in keeping with AIPS++
philosophy, the reader is referred to the AIPS++ header files for more
information.

We assume that the reader is familiar with the terminology,
notation and results from the above mentioned references.

\section{Goals}

The goals for the design of the calibration and imaging capabilities were described
by Cornwell and Wieringa (1996). Here we describe the additional goals for
cross-calibration:
\begin{itemize}
\item Describe calibration information in the form of {\tt MeasurementComponents},
\item Allow very general selection of data for calibration determination
(e.g. ``All 6cm observations where wind velocity$\leq$3 m/s''),
\item Allow direct manipulation of calibration tables,
\item Allow iterative calibration,
\item Allow incremental calibration,
\item Allow construction of {\tt MeasurementComponents} from externally-derived
data or from calibration solutions,
\item Allow specification of source structure via SkyModels,
\item Keep all data in one place.
\end{itemize}

\section{Some comments on capabilities available inside AIPS++}

\begin{description}
\item[Tables] All data, original and calibration, is placed in the
Table system, where it is accessible and modifiable by a number of
``free-form'' mechanisms, the principle being glish and the tablebrowser.
\item[The MeasurementSet] is of the form one main Table, plus many
subsiduary Tables (e.g. ANTENNA, FEED, ARRAY). The main table contains data, 
coordinate information such as time and uvw, and keys into the sub-Tables (see Wieringa 
and Cornwell, 1996).
\begin{description}
\item[keys] ANTENNA1, ANTENNA2, ARRAY\_ID, CORRELATOR\_ID, FEED1, FEED2, FIELD\_ID,
OBSERVATION\_ID, PULSAR\_ID, SOURCE\_ID, SPECTRAL\_WINDOW\_ID
\item[coordinates] SCAN\_NUMBER, UVW, TIME, EXPOSURE, INTERVAL, PULSAR\_BIN
\item[data] DATA, SIGMA, WEIGHT, WEIGHT\_SPECTRUM, FLAG, FLAG\_ROW
\end{description}

\item[Selection] of data from a MS leads to another reference MS with
little overhead. The selection can be by one or more of:
\begin{itemize}
\item data (e.g. ``all amplitudes $\geq$ 5 Jy''), 
\item coordinates (e.g. ``all times in the following range''), 
\item keys (e.g. ``SPECTRAL\_WINDOW\_ID=1 or 4''), 
\item properties determined by lookup in a subtable (e.g. ``wind velocity
$\leq$ 3 m/s''), 
\end{itemize}
\item[The {\tt MeasurementEquation}] is the fundamental mechanism used to
describe calibration and imaging in terms of {\tt
MeasurementComponents}. These describe some aspect of the measurement
process, such as antenna-based gain, while not prescribing how the
calculation of actual values is to be performed, for instance in the
case of antenna phases via a table calculated from a priori values or
via a phase-screen model.
\end{description}

\section{An approach to cross-calibration}

Our preferred approach to calibration uses the reference table or MS
concept to keep both the original data together in one MS and to provide
calibrated views of that data via reference tables.
\begin{itemize}
\item The {\tt MeasurementSet} $M$ is subject to some
selection, $S_{cal}$, to produce a logical subset MS $M(S_{cal})$, the
calibrator MS.  The selection used is of interest in two ways: first
to ensure that the correction is applied to the relevant data
(e.g. 6cm to 6cm observations), and second as information to the user
(e.g.  ``wind-velocity $\leq$ 3 m/s'').
\item For each such calibrator MS, a SkyModel is used to predict the corresponding coherences.
\item For each such calibrator MS, one will want to perform a solution
for one or more {\tt MeasurementComponents} (e.g. {\tt GJones} and
{\tt DJones}). 
\item Each {\tt MeasurementComponent} is then stored as a Table in the calibrator
MS. 
\item The {\tt MeasurementSet} $M$ is subject to some
selection, $S_{source}$, to produce a logical subset MS $M(S_{source})$, the
source MS.
\item {\tt MeasurementComponents} are then explicitly copied from calibrator to
source MS, either by the user or by some simple program.
\item After copying, a {\tt MeasurementComponent} can be subject to a number of
operations:
\begin{itemize}
\item Re-interpolation in some coordinate such as time or some derived
quantity such as antenna elevation or airmass,
\item Editing by hand or by algorithm,
\item Boot-strapping of some form (e.g. determination of the flux of
some sources from a primary calibrater),
\item Recasting to another form (e.g. antenna-based phases to
position-based phase screen).
\end{itemize}
All of these are probably best thought of as construction of a new {\tt
MeasurementComponent} that replaces the initial one. 
\end{itemize}

The net result is that one main MS and a number of reference MSes pointing to
it and containing the relevant {\tt MeasurementComponents}. This scheme
has a number of advantages:
\begin{enumerate}
\item (Relative) simplicity. We consider this to be very important,
\item The {\tt MeasurementComponents} in the source MSes exist independently of the
calibrator MSes and can be edited or munged as one chooses,
\item Self-calibration is easily accomodated by simply updating the 
{\tt MeasurementComponents} in the source MSes.
\end{enumerate}

An alternate but more complicated approach is to defer the connection
between sources and calibrators until the correction of data is
actually performed. In this scheme, each reference MS is assigned a
unique integer, CG, and an MS column ``CALIBRATION\_GROUP'' is set to
that integer. The links between calibrator and source are then made
via this index. The data for which these corrections are to be applied
are selected via a selection $S_{source}$ to form a target MS,
$M(S_{source})$. The column CALIBRATION\_GROUP of this MS, is set to
CG.  We discarded this type of deferred selection of {\tt
MeasurementComponents} as being overly complex, and not allowing the
user to munge the {\tt MeasurementComponents} ``by hand''.


\section*{References}

\noindent Cornwell, T.J., 1995, AIPS++ Implementation Note 183.

\noindent Cornwell, T.J. and Wieringa, M.H., 1996, AIPS++ Implementation Note 189.

\noindent Hamaker, J.P., Bregman, J.D., and Sault, R.J., 1995,
{\em Understanding radio polarimetry: I Mathematical foundations},
submitted to A\&A.

\noindent Noordam, J., 1995, AIPS++ Implementation Note 185.

\noindent Wieringa, M.H., and Cornwell, T.J., 1996, AIPS++ Implementation Note 191.

\end{document}
