%% VLBI++103
%%----------------------------------------------------------------------------
\documentclass[11pt]{article}
\usepackage{html, epsfig, longtable, aips2help}
\input{aips2defs.tex}
%%-----------------------------------------------------------------------------

\begin{document}

\title{AIPS++ DEVELOPMENT PLAN: Release 1.5}
  

\author{Athol Kemball and Tim Cornwell (eds.)\\
   NRAO}

\date{15 Nov, 2000}

\maketitle
%%---------------------------------------------------------------------------
\begin{htmlonly}
\htmladdnormallink{A postscript version of this note is available (47kB).}{../239.ps.gz}
\end{htmlonly}

\tableofcontents
      
\newcommand{\remark}[1]{{\small\begin{itemize}\em \item #1\end{itemize}}}
\newcommand{\prevdocURL}{http://aips2.nrao.edu/aips++/docs/notes/226/226.html}
\newcommand{\thisdocURL}{http://aips2.nrao.edu/aips++/docs/notes/234/234.html}
\newcommand{\systemplanURL}{../220/220.html}
\newcommand{\searchURL}{../../html/aips2search.html}
\newcommand{\aipsviewURL}{../../user/aipsview/aipsview.html}
\newcommand{\GlishmanualURL}{../../reference/Glish/Glish.html}
\newcommand{\GlishtutorialURL}{../../notes/195/195.html}
\newcommand{\synthesisURL}{../../user/synthesis/synthesis.html}
\newcommand{\gsURL}{../../user/gettingstarted/gettingstarted.html}
\newcommand{\userrefmanualURL}{../../user/Refman/Refman.html}
\newcommand{\specsURL}{../../specs/specs.html}
\newcommand{\betaURL}{../beta/beta.html}
\newcommand{\consortiumURL}{../consortium.html}
\externallabels{../../reference/Glish}{../../reference/Glish/labels.pl}
\externallabels{../../user/Utility}{../../user/Utility/labels.pl}
\externallabels{../../user/General}{../../user/General/labels.pl}
\externallabels{../../user/Display}{../../user/Display/labels.pl}
\externallabels{../../user/Dish}{../../user/Dish/labels.pl}
\externallabels{../../user/Synthesis}{../../user/Synthesis/labels.pl}
\externallabels{../../user/ATNF}{../../user/ATNF/labels.pl}
\externallabels{../../user/NFRA}{../../user/NFRA/labels.pl}
\externallabels{../../user/NRAO}{../../user/NRAO/labels.pl}
%

\section{\underbar{Purpose}}

The purpose of this document is to define the development plan for
AIPS++ release v 1.5, currently scheduled for April 30, 2001.

\section{\underbar{Release priorities}}

The highest priority for the project at this time is scientific
completeness, and planning for release v1.5 has been undertaken in
keeping with this objective. Maintaining a sound infrastructure in the
project is important however, and infrastructure work has been
scheduled for this cycle carefully; primarily in areas which are vital
for the long-term vitality of the project, or which are in the
critical path for application development.

\section{\underbar{Introduction}}

Development priorities and targets are listed separately for each
major development area in AIPS++. These targets cover only the
development cycle through April 2001, and do not include longer-term
items, which are tracked separately for consideration in subsequent
development cycles (docs/project/devplan\_1.6.ps.gz). There are 21
weeks between the week of November 20, 2000, when interim targets for
v1.5 were first set, and the expected code-freeze on 13 April 2001, but
deductions for defect correction (30\%; 1.5 days per week), user
support (up to 20\%), and where applicable, science time (25\%), have
been made. The exact numbers vary by developer and institution, and
the best estimate has been made in each case. Variability in these
estimates is accommodated by the inclusion of medium prioirty targets,
which may need to be carried over into the next cycle. Each target is
assigned a priority as high ({\bf H}), which implies that the target
is expected to be completed this cycle, and medium ({\bf M}), which is
work to be done on an as-available basis. Note also, that not all
developers are available to the project on a full-time basis, and have
reduced commitments as a result.

\vskip 1cm

\begin{tabular}{|l|l|l|}\hline
{\bf Abbr.}&{\bf Developer}&{\bf Affiliation}\\ \hline\hline
AK & Athol Kemball & NRAO\\
AS & Anuj Sarma & NCSA\\
BG & Bob Garwood & NRAO\\
BM & Barry Maguire & NRAL\\
DK & David King & NRAO\\
DM & Dave Mehringer & NCSA\\
DS & Darrell Schiebel & NRAO\\
GvD & Ger van Diepen & NFRA\\
GM & George Moellenbrock & NRAO\\
HR & Harold Ravlin & NCSA\\
JB & Jim Braatz & NRAO\\
JM & Joe McMullin & NRAO\\
KG & Kumar Golap & NRAO\\
MH & Mark Holdaway & NRAO\\
MM & Malte Marquarding & ATNF \\
MW & Mark Wieringa & ATNF \\
NK & Neil Killeen & ATNF\\
PC & Paulo Cortes & NCSA\\
PT & Peter Teuben & BIMA \\
RM & Ralph Marson & NRAO\\
RP & Ray Plante & NCSA\\
OS & Oleg Smirnov & NFRA\\
TC & Tim Cornwell & NRAO\\
WB & Wim Brouw & ATNF\\
WY & Wes Young & NRAO\\
\hline
\end{tabular}

%%----------------------------------------------------------------------

\section{\underbar{Single-dish}}

\subsection{\underbar{Priorities}}

\begin{description}

\item[Support GBT commissioning] The continued development of tools
required for GBT commissioning remains a high priority during this
development cycle.

\item[Calibration and imaging capabilities] The provision of
calibration and imaging capabilities which are integrated with
the synthesis infrastructure is a high priority.

\item[Responding to user feedback] This is a high priority as many new
users will be introduced to the single-dish software this development
cycle through GBT commissioning.

\end{description}


\subsection{\underbar{Targets}}

\begin{description}

\item[Consistent use of flags, weights throughout dish] ({\bf JM, H, 3 wk}).

\item[SD imaging using imager] Extend and test the current SD imaging
capabilities in imager. ({\bf JM, H, 2 wk, TC, H, 3 wk}).

\item[Spectrometer backend support] ({\bf BG, H, 2 wk}).

\item[IF and LO1 support] ({\bf BG, H, 2 wk}).

\item[Initial SD calibration using calibrater] Initial calibration
examples using a common infrastructure with synthesis. ({\bf JM, H, 3
wk, AK, H, 2 wk}).

\item[GBT commissioning assistance] Continued assistance in
GBT commissioning; total time listed here. ({\bf JB, H, 7 wk, BG, H, 4 wk}).

\item[dish test scripts] Automated test scripts for dish. ({\bf JM, H, 2 wk}).

\item[Improved single-dish documentation] ({\bf BG, H, 2 wk}).

\item[Replace sdimager] ({\bf JM, M, 1 wk}).

\item[Bad pixel replacement] Replace bad pixels by various means ({\bf
BG, M, 2wk}).

\end{description}

%%----------------------------------------------------------------------

\section{\underbar{Synthesis}}

\subsection{\underbar{Priorities}}

\begin{description}

\item[Scientific completeness] This remains the primary driver in
synthesis development. Significant synthesis capabilities exist within
the package at present, and are being more widely used in the
scientific community with each release. This process can be aided by
continued improvement in vertical integration of synthesis
applications, and in the addition of scientific completeness beyond
current capabilities.

\item[Time-critical local priorities] It is important that the project
meet time-critical targets required for successful AIPS++ use at
consortia sites, particularly where the use of AIPS++ is closely
integrated into the critical path required for instrument use or
operation.

\item[Basic automated imaging] Significant work in this area took
place in v1.4, but needs to continue in this development cycle to
include the highest level tools in this area.

\item[Automated testing] Expanding the automated testing of synthesis
capabilities using simulated data is a priority during this cycle,
both at the C++ and Glish level. This has benefits in speeding up
release testing, and in daily and weekly checks of system integrity
for end-users and user test groups.

\end{description}

\subsection{\underbar{Targets}}

\begin{description}

\item[BIMA filler migration] Migrate the BIMA filler to MS v2. ({\bf
PT, H, 3 wk}).

\item[MS v2 FITS definition] Complete definition of MS v2 in an
external FITS binary table format. ({\bf GM, H, 2 wk}).

\item[Additional VLA filler features] Add: (i) shadowing; (ii) Tsys
and weighting; (iii) complete output filtering; (iv) new weights
scheme currently adopted by FILLM. ({\bf RM, H, 4 wk}).

\item[MS data volume control] Implement initial solution to the MS data volume
problem caused by the MODEL\_DATA, CORRECTED\_DATA and IMAGING\_WEIGHT
scratch columns. ({\bf AK, H, 1 wk}).

\item[Automated editing] Complete initial automated editing
algorithms, including spectral rejection. ({\bf OS, H, 8 wk}).

\item[Investigate coupled solver for ATCA] Complete investigate a coupled
solver for ATCA data. Includes a comparison with MIRIAD. ({\bf AK, H, 2 wk}).

\item[Complete integration of calibration table re-gridding] Complete
integration of existing table smoothing and re-gridding tool. ({\bf
AK, H, 2 wk, HR, H, 2 wk, DM, H, 1 wk}).

\item[Ephemeris-based planet flux density calibration] Enable flux
density calibration based on planet observations, as currently done in
Miriad. ({\bf RP, M, 4wk}).

\item[Holography support in imager] Complete general holography support
in imager, excluding near-field corrections. ({\bf JB, H, 4 wk}).

\item[Optimal-sized FFT's for mosaicing] Complete the implementation
of optimal-sized FFT's for mosaiced imaging. ({\bf MH, H, 2 wk}).

\item[Simulator error models] Expand the analytic error models in the
simulator for use in correctness testing. Also, complete initial
support of pointing errors({\bf MH, H, 3 wk}).

\item[Automated imaging prototype] Provide an initial automated
imaging utility, automap.g. ({\bf AK, H, 2 wk}).

\item[Pipeline scientific requirements] Complete the scientific
requirements document for common pipeline architectures. ({\bf AK, H, 1 wk}).

\item[New Getting Results chapters] Add new Getting Results chapters
for synthesis reduction for specific instruments, including VLA and
BIMA. ({\bf AK H, 1 wk, RP, H, 1 wk, DM, H, 1 wk}).

\item[Synthesis tests] Expand test scripts for all consortium
connected-element instruments (VLA, ATNF).  ({\bf GM, H, 2 wk, DM, H, 6
wk, RP, H, 2 wk}).

\item[Errors in componentlist tables] Add support for simple errors
(one per existing parameter) in componentlist tables. ({\bf RM, H, 1
wk}).

\item[Ionosphere support in calibrater] Complete FJones support in
calibrater, including: i) finalize FVisJones classes; ii) test
PIM-based corrections with WSRT data; and, iii) add secondary PIM
corrections. ({\bf OS, H, 6 wk}).

\item[imager tests against simulated data] Add imager tests against
simulated data for all imaging modes to imagertest, and assay. ({\bf
GM, H, 3 wk}).

\item[VLBI FITS-IDI filler] Finalize existing FITS IDI filler. ({\bf
BM, H, 6 wk}).

\item[Initial fringe-fitter] Complete initial single-band, coherent
fringe fitter. ({\bf GM, H, 4 wk}).

\item[Mosaic and wide-field demonstrations] Complete demonstration
scripts for mosaic, and wide-field imaging. ({\bf MH, H, 2 wk}).

\item[Improved total-power support in mosaicing] ({\bf MH, H, 4wk}).

\item[calibrater infrastructure] Complete calibrater infrastructure
development in interpolation and parametrized solvers ({\bf AK, H, 4 wk}).

\item[Custom synthesis GUI] Initial custom synthesis GUI. Includes
variants for individual instruments ({\bf AK, H, 2 wk, DM, M, 2 wk}).

\item[ATCA filler enhancements] i) birdies; ii) syscal flagging; iii)
Hanning smoothing; iv) shadowing; v) documentation ({\bf MW, M, 3
wk}).

\end{description}

%%----------------------------------------------------------------------

\section{\underbar{Applications integration}}

This includes development aimed at improving the overall integration
of applications in AIPS++ in a common framework, using common
interfaces and services.

\subsection{\underbar{Targets}}

\begin{description}

\item[Style guide for Refman] ({\bf NK, H, 1 wk}).

\item[Revise and update programmer doc] Update programmer
documentation. Includes exception-handling guidelines and a review of
Glish coding rules and templates ({\bf AK, M, 1 wk}).

\end{description}

%%----------------------------------------------------------------------

\section{\underbar{Build and code distribution system}}

This covers all system work, excluding maintenance and support of the
basic library.

\subsection{\underbar{Priorities}}

\begin{description}

\item[Stability] A fundamental requirement in this cycle is the
continued provision of a stable build to maximize application
development and testing efficiency. The project is now operational and
build failures have a significant impact on applications development
efficiency and the ease of user support.

\item[Initial developer support] An initial group of external
developers is starting to use the system for exploratory
development. This group can be supported in the most cost-effective
manner by starting to provide formal support for external development
on very restricted platforms in advance of the full developer's
release, which is planned for 2001.

\end{description}

\subsection{\underbar{Targets}}

\begin{description}

\item[Complete initial developer's release] Document the makedefs, and
provide an installation FAQ for an initial developer's release,
targeted at RedHat and SuSe Linux systems of restricted version
numbers. Test the resulting CDs. ({\bf DS, H, 2 wk}).

\item[Update the project compiler] Evaluate gcc 2.95.2 as the project
compiler; adopt if review and associated change proposal are
accepted. ({\bf AK, H, 2 wk}).

\end{description}

%%----------------------------------------------------------------------

\section{\underbar{Glish}}

\subsection{\underbar{Priorities}}

\begin{description} 

\item[Maintenance] The work on Glish is in a stage of consolidation
and maintenance. In general, Glish's capabilities are sufficient both
for developers to implement higher-level applications and for users to
explore their data. The language is also at about the right level; it
is capable enough for complex scripts, while still being approachable
by general users. 

\end{description}

\subsection{\underbar{Targets}}

\begin{description}

\item[Understand speed of TCL/Tk and Glish/Tk] There is currently a
perceived performance problem in Glish/Tk, which is especially
pronounced when drawing large GUIs. The specific origin is
unknown. This target involves determining the specific cause by direct
comparison with TCL/Tk, including investigating what fraction of the
speed differential is caused by byte-compiling. If simple remedies are
apparent short of multi-threading, this target includes their
implementation. ({\bf DS, H, 4 wk})

\end{description}

%%----------------------------------------------------------------------

\section{\underbar{User interface and tasking}}

\subsection{\underbar{Priorities}}

\begin{description}

\item[Consolidation and evaluation] Views on user interfaces are
inherently subjective; a wider pool of scientific users needs to use
and comment on the current automated GUI system, as implemented in
toolmanager, before undertaking a second revision of the user
interface. This feedback needs to be actively sought, and reviewed
during this cycle. This factor, combined with the limited available
resources for toolmanager development, argues for consolidation only
during this development cycle.

\item[New technology evaluation] It is appropriate at this point to
evaluate new technology for the Tasking and UI systems.

\end{description}

\subsection{\underbar{Targets}}

\begin{description}

\item[toolmanager] Maintenance of toolmanager, and addition of new
capabilities in the area of integration and scientific
usability. Consolidation of related code in identified areas, and
speed improvements ({\bf RM, H, 6 wk}).

\item[Requirements document for revised Tasking/UI system] Draft
requirements for an upgrade of the current Tasking and User Interface
system. ({\bf AK, H, 1 wk})

\item[Implementation proposal/prototypes for revised Tasking/UI] Draw up an
implementation proposal and schedule for moving to a revised
Tasking/UI system, including proposed middleware, such as CORBA, and
the associated Glish interface. Includes simple prototyping and
related evaluation. ({\bf DS, H, 5 wk}).

\end{description}

%%----------------------------------------------------------------------

\section{\underbar{Basic library}}

\subsection{\underbar{Priorities}}

\begin{description}

\item[Consolidation] The basic library, which covers all fundamental
infrastructure, is mostly complete. Development in this cycle has been
scheduled in the area of library maintenance, and in the provision of
limited new capabilities required for application development.

\end{description}

\subsection{\underbar{Targets}}

\begin{description}

\item[Large-file support in Tables] Add large-file support in the
Table system. Includes evaluation of trial LFS available for Linux
({\bf GvD, H, 3 wk}).

\item[Sub-table locking options] Add expanded sub-table locking options
({\bf GvD, H, 0.5 wk}).

\item[Arrays in TaQL] Support arrays in all TaQL functions ({\bf GvD,
H, 1 wk}).

\item[Column removal in TSM] Allow column removal in TSM ({\bf GvD, H, 1 wk}).

\item[Deep copy and actualDesc()] Provide additional information on
the storage manager in use and the actual table description; allow
deep copy of tables. ({\bf GvD, H, 1 wk}).

\item[Time comparison function in TaQL] Add a time comparison function
to TaQL, to specifications provided by synthesis. {\bf GvD, H, 1 wk}).

\item[Exception-handling clean-up] Remove CanDelete and Cleanup from
the library ({\bf DS, H, 0.25 wk}).

\item[File descriptor limit] Change aipsinit scripts to set
ulimit/limit on maximum number of file descriptors appropriately ({\bf
DS, H, 0.25 wk}).

\item[complex, string, and iostream] Move to standard library
compliance in these areas. Build stability to be maintained in this
process. ({\bf WB, H, 2 wk}).

\item[Measures and Quanta] i) tests of baseline, antenna positions and
uvw calculations; ii) arrays of measures in Glish; and, iii) quanta
improvements per defects and release testing. ({\bf WB, H, 4 wk}).

\item[Fitting] i) Image-fitting support; ii) requirements document for
component model changes for uv-fitting. ({\bf WB, H, 3 wk}).

\item[Measures chapter for Getting Results] Chapter in Getting Results
describing end-user Glish use of Measures. ({\bf RP, M, 2 wk}).

\end{description}

%%----------------------------------------------------------------------

\section{\underbar{Images}}

\subsection{\underbar{Targets}}

\begin{description}



\item[Imagefitter] i) add support for error propagation to component
models; ii) corrections for spatial correlation. ({\bf NK, H, 3 wk}).

\item[Stokes coordinate handling] Fix Stokes coordinate problem
relating to the use of the reference value ({\bf NK, H, 0.5 wk}).

\item[Relative coordinates] Support relative coordinates ({\bf NK, H, 1 wk}).

\item[DSS FITS images] Handle DSS FITS images correctly ({\bf NK, H, 1 wk}).

\item[Extension regions] Improve extension region handling ({\bf NK,
H, 2 wk, GvD, H, 1 wk}).

\item[Spectral fitter] Fit spectra in images ({\bf NK, H, 2 wk}).

\item[Coordinates review] Finish coordinates review ({\bf NK, H, 0.5 wk}).

\item[Prepare LatticeStatistics and LatticeHistogram for review]
Prepare these classes for code review. ({\bf NK, H, 1 wk}).

\item[Vector DisplayData] Implement an initial display data class for
plotting vectors. ({\bf NK, H, 2 wk}).

\end{description}

%%----------------------------------------------------------------------

\section{\underbar{Visualization}}

This includes work in the Display Library (DL), and in applications
using this library.

\subsection{\underbar{Priorities}}

\begin{description}

\item[Higher-level image visualization] The image visualization
interface is sufficiently advanced to allow the development of initial
high-level image applications.

\item[Provision of uv-visualization services] The DL has provided
excellent support of image visualization, and continuation of the
effort to expand this capability to uv-data visualization is a high
priority.

\end{description}

\subsection{\underbar{Targets}}

\begin{description}

\item[kslice3d equivalent] Design and implement a kslice3d equivalent
({\bf MM, H, 4 wk}).

\item[Color-wedge DD] ({\bf MM, H, 2 wk}).

\item[viewer support of PanelDisplay and MWCAnimator] Move from WC in
viewer to PanelDisplay and add panelling to viewer. Add MWCAnimator to
viewer ({\bf MM, H, 3 wk}).

\item[Relative coordinate support] ({\bf MM, H, 1 wk}).

\item[Improve SkyCatOverlay] Use TableMeasures. ({\bf MM, H, 1 wk}).

\item[Datamanager improvements] FileBrowser and drag-and-drop
capabilities ({\bf MM, H, 2 wk}).

\item[Implement DisplayShapes as base class] Re-implement rectangle
DD, finish ellipse, and add polygon, line and crosshair ({\bf MM, M, 2
wk}).

\item[Vertical integration of uv-DD] Make uv-DD visible in the viewer,
starting with TableAsRaster ({\bf DK, H, 3 wk}).

\item[Completion of MS uv-DD] Completion of MSAsXY, MSAsRaster and
MSAsContour ({\bf DK, H, 3 wk}).

\item[Editing EH] Basic editing in uv-DD using DL EH ({\bf DK, H, 3 wk}).

\item[Coordinates on WC] Implementation proposal for coordinates on
the WC ({\bf DK, H, 2 wk}).

\item[Initial interactive editor using TB display data] Complete the
first TVFLG-like interactive editor using the TB display data. ({\bf DK,
H, 3 wk}).

\item[OpenGL or vtk PixelCanvas] Investigate OpenGL or vtk PixelCanvas
({\bf HR, M, 6 wk}).

\end{description}

%%----------------------------------------------------------------------


\section{\underbar{Parallelization and high-performance computing}}

\subsection{\underbar{Priorities}}

\begin{description}

\item[Scientific application] The parallelization effort needs
to demonstrate a scientifically useful capability to address the most
challenging problems in radio astronomy where supercomputer resources
are required. These include wide-field imaging problems at low
observing frequencies, mosaicing and the largest VLBI observations,
amongst others. In addition, this also includes new algorithms which
have not been widely used to date due to limited computing resources.

\item[Parallelization infrastructure] A central goal of the
parallelization effort is to ensure that infrastructure is developed
within the AIPS++ system as a whole to support parallel and
distributed computing without expensive ad hoc modifications. This
requires that the parallelization infrastructure be compatible with
the overall project design, and also that the mainstream project
development consider parallelization when implementing algorithms. It
is also imperative that the parallelization capabilities be presented
using the same user interface as the conventional package.

\item[High-performance computing in AIPS++] The parallelization
effort has a strong vested interest in the serial performance of
AIPS++ for problems of the largest size, which are defined to be those
with exceptional I/O, memory or CPU requirements. It is considered the
responsibility of this group to profile the serial performance in
these specialized cases, make any changes required to support these
large problem sizes, and optimize overall serial performance in these
cases.

\end{description}

\subsection{\underbar{Targets}}

\begin{description}

\item[Cluster Linux and IRIX build maintenance] Continued maintenance of the
existing NCSA/NRAO builds under IRIX and on the AHPCC Linux
cluster. ({\bf WY, H, 3 wk, DM, H, 3 wk}).

\item[Key project processing] Processing of five key projects
(including at least one each of mosaiced, wide-field or large spectral
line). Candidates include the existing M33 dataset (Westpfahl), TXCam
(Kemball), and a selection of low-frequency VLA projects in
A-configuration. Generation of user liaison documentation and user
support at NCSA. ({\bf RP, H, 2 wk, AS, H, 2 wk, WY, H, 2 wk}).

\item[Complete multi-field parallelization] Complete implementation of
a prototype parallelization for mosaiced or wide-field
imaging. Candidates include field- or facet-based gridding, model
prediction or residual image computation. ({\bf KG, H, 4 wk}).

\item[Parallelization of other deconvolution methods] Extend the Clark
CLEAN parallelization to other deconvolution algorithms. ({\bf KG, H,
4 wk}).

\item[Initial parallel I/O implementation] Implementation of
multi-process I/O on the same file, multi-iterator access in a single
process, and ROMIO asynchronous I/O using MPI-2. Evaluation of
performance in these cases. ({\bf WY, H, 4 wk}).

\item[Complete NT migration document] Combine work by P. Cortes in NT
migration to formalize incorporation of these changes in the main code
distribution ({\bf WY, H, 1 wk, PC, H, 3 wk}).

\item[Modify test suite for large problem size] Modify the current
bigimagertest to use simulated data. ({\bf DM, H, 1 wk}).

\item[Cluster parallelization] Expand evaluation of parallelization
methods specific to cluster architectures, with a specific focus on
Linux systems. Port AIPS++ to IA64 when available. ({\bf PC, H, 7 wk})

\end{description}

%%----------------------------------------------------------------------

\end{document}


