\documentclass{article}
\include{aips2defs}
\usepackage{html,epsf}
\setlength{\textwidth}{6.5in}
\setlength{\evensidemargin}{0.0in}
\setlength{\oddsidemargin}{0.0in}
\parskip=5truept
\newcommand{\nchan}{N_{chan}}
\newcommand{\nrecep}{N_{receptors}}
\newcommand{\nspw}{N_{spw}}
\newcommand{\njones}{N_{jones}}
\newcommand{\ndir}{N_{dir}}
\newcommand{\defline}[1]{\cline{1-5}
\multicolumn{5}{|l|}{#1} \\
\cline{1-5}}
\newcommand{\definetable}[2]
{
	\vfill\newpage
	\subsection{#1}
        \vspace{0.15in}
        \small
	\begin{tabular}{|l|p{1.25in}|l|p{.9in}|p{1.4in}|}
	\hline
	\multicolumn{5}{|c|}{\bf #1}\\ 
	\cline{1-5}
        \multicolumn{1}{|c|}{Name}&\multicolumn{1}{|c|}{Format}&
        \multicolumn{1}{|c|}{Units}&\multicolumn{1}{|c|}{Measure}&
        \multicolumn{1}{|c|}{Comments}\\
        \cline{1-5}
        #2
        \hline
	\end{tabular}
}
\begin{document}

\title{Calibration table definition version 2.0}
\author{A.J. Kemball (NRAO)}
\date{January 01, 2001}
\maketitle

\begin{htmlonly}
\htmladdnormallink{A postscript version of this note is available 
(59kB).}{../240.ps.gz}
\end{htmlonly}

\tableofcontents 
\section{Summary} 

This note describes the format of calibration tables in \aipspp. The
calibration table format is currently at revision v2.0. 

\section{Introduction}

The calibration and imaging formalism used in \aipspp is based closely
on the generic instrument model of Hamaker, Bregman and Sault
(1996). This formalism, referred to as the Measurement Equation (ME),
is instrument-independent and represents calibration effects in both
the visibility and image plane as Jones matrices, with arbitrary
parametrization and polarization basis. It's use in \aipspp is
described by Cornwell (1995) and Noordam (1995), and references
therein.

The calibration table format used in \aipspp was designed to support
the fundamental principles on which the ME formalism is based. The
definition of a standard calibration table format allows calibration
components to be stored permanently for subsequent retrieval,
application or modification. The current calibration table format
revision level is v2.0.

The underlying principles on which the calibration table data format
is based are described below.

\begin{itemize}

\item{{\bf Data representation:} \aipspp calibration tables are stored
as \aipspp {\it Tables}, in keeping with the general rule in \aipspp
for all data visible to end users. As such, there is full user access
to all data stored in a calibration table. These data can be
retrieved, inspected and modified in the general user interface. The
implementation of the calibration tables also uses other common {\it
Tables} infrastructure such as {\it TableMeasures}.}

\item{{\bf ME formalism:} the calibration table format was chosen to
support the key assumptions of the ME formalism, including support for:
i) arbitrary polarization bases; ii) arbitrary Jones matrix
parametrization; and, iii) generic, instrument-independent
representation of calibration components.}

\item{{\bf Unified calibration:} the calibration table format has been
designed to support both single-dish and synthesis calibration,
antenna- and baseline-based Jones matrices, as well as
visibility-plane and image-plane calibration components in a common
format, with specialization where appropriate. This was chosen to
maximize the re-use of calibration information and allow it to be
transferred between different observing contexts.}

\item{{\bf MS integration:} the calibration table format is fully
integrated with the MeasurementSet (MS) data format, supports all data
representations possible in the MS, and re-uses all applicable column
and keyword names for common underlying physical concepts. A
calibration table can be attached to one or more MeasurementSets, but
exists as a stand-alone table with a user-specified name. This
supports both single-project and multi-project calibration schemes.}

\item{{\bf History tracking:} the calibration table format supports
full tracking of the calibration history attached to each calibration
solution or step in the calibration process.}

\end{itemize}

\section{Calibration table format}

This section describes and defines all fields in the calibration table
format. At the highest level, the calibration table consists of a main
table, CAL\_MAIN, and two sub-tables stored as table keywords
CAL\_DESC and CAL\_HISTORY. The CAL\_DESC sub-table describes the
shape and coordinates of the array columns in the main calibration
table. CAL\_HISTORY references the associated calibration history for
each row in the main table. All tables are described separately below.

\subsection{CAL\_MAIN}

The ME contains several different Jones matrix calibration components
which are enumerated in the references listed above. In general,
the Jones matrix calibration components fall into the following broad
categories:

\begin{description}

\item{{\bf Antenna-based, visibility-plane, non-solvable:} i) {\bf P}
- parallactic angle; and, ii) {\bf C} - polarization configuration.}

\item{{\bf Antenna-based, visibility-plane, solvable:} i) {\bf T} -
atmospheric correction; ii) {\bf G} - electronic gain; iii) {\bf D} -
instrumental polarization; iv) {\bf B} - bandpass; and v) {\bf F} -
ionosphere correction.}

\item{{\bf Interferometer-based, visibility-plane, solvable:} The only
supported interferometer-based correction at present is a
multiplicative, solvable term ({\bf M}). An additive term is also
allowed by the ME.}

\item{{\bf Antenna-based, image-plane, non-solvable:} i) {\bf P} -
parallactic angle.}

\item{{\bf Antenna-based, image-plane, solvable:} i) {\bf T} -
atmosphere correction; ii) {\bf D} - instrumental polarization; iii)
{\bf F} - ionosphere correction; and iv) {\bf E} - voltage pattern.}

\end{description}

Note that the same Jones matrix can appear as both a visibility-plane
and image-plane correction in the ME, as appropriate. In addition, an
individual Jones matrix component may either be discretely sampled or
modeled using arbitrary parameters.

Jones matrices of different types (time-variable, solvable,
baseline-based, antenna-based and parametrized) share as many common
sections in the CAL\_MAIN table as appropriate, thus representing the
inheritance and specialization inherent to the family of Jones matrix
types. This inheritance relationship is illustrated by the {\it
VisJones} and {\it SkyJones} class families in the \aipspp library.
Each calibration table holds only Jones calibration matrices of one
type (e.g. {\bf GJones}, or {\bf PJones} etc.).

The key sections in the CAL\_MAIN table are as follows:

\begin{itemize}

\item{{\bf Primary MS indices:} The first section contains the MS
indices representing the MS data from which the calibration solution
was derived or which otherwise label the calibration data. Fundamental
amongst these are the time and interval of each row. Indices over
which the MS data were averaged prior to the solution, or which are
not applicable, are set to -1.}

\item{{\bf Secondary MS fields:} The primary MS indices can be used to
retrieve any secondary MS information by direct lookup in the
associated MS. However, a limited subset of MS information is stored
directly in the calibration table, to minimize the need for lookup in
the MS for the most common secondary information. This breaks database
normalization, but as is commonly the case, is warranted when dictated
by efficiency. This also allows the calibration table to be used in
many cases when the associated MS is not available. Where these column
names are not unique within the MS as a whole, they are prefixed by
the associated MS subtable name (e.g. SOURCE\_CODE).}

\item{{\bf Gain values:} The calibration gain factors, represented as
sampled complex $(1 x 1)$, $(2 x 2)$ or $(4 x 4)$ Jones matrices, in
array format over optional axes of spectral window identifier,
frequency channel and sky coordinates, are stored in a GAIN
column. This column is optional for parametrized calibration
components.}

\item{{\bf Reference frame:} A section is provided to represent the
reference antenna, feed, receptor, frequency and direction for
calibration parameters, where appropriate.}

\item{{\bf Solution statistics:} A section is defined to contain the
statistical properties of each calibration solution. This includes
boolean flags for the calibration gain arrays and the fit and fit
weight, both per array element and for the solution interval as a
whole.}

\item{{\bf Jones matrix parametrization:} Customized columns required
to represent parametrized Jones matrices are defined in a separate
section. These are different for each parametrized type.}

\item{{\bf Sub-table pointers:} Indices per CAL\_MAIN row, into
the CAL\_DEC and CAL\_HISTORY sub-tables, are defined separately.}

\end{itemize}

\subsection{CAL\_DESC}

The CAL\_DESC sub-table defines the dimensions of the array-based
columns in CAL\_MAIN, and specifies the coordinates of the array axes.
These coordinates include frequency, receptor polarization, and for
image-plane components, directions and regions in the image-plane.

The frequency labeling of discretely sampled calibration gain values,
as well as parametrized calibration models, may differ from the
frequency labeling in the associated MS, if calibration is derived by
aggregating or sub-dividing the MS spectral windows. Hence, the
spectral windows and frequency channels along the calibration array
axes, as described in CAL\_DESC, may differ from the MS itself. Where
there is a direct mapping to MS spectral windows and channel ranges,
this is recorded. 

The polarization axis is defined in terms of a list of receptor
polarizations. As an enumerated coordinate, these overlap with those
used in the associated MS.

For parametrized calibration components, the coordinate axes in
CAL\_DESC define the coordinate ranges over which the parameters are
valid, and also the dimensions of the model parameters stored as
arrays. In this case, the frequency and direction coordinates also
define the units and frame for the parameters.

The image-plane direction array axis is defined by specifying an array
or directions and associated regions about each direction in the image
plane. This allows arbitrary discrete sampling of image-plane Jones
calibration matrices, as well as arbitrary parameter domains for
image-plane calibration models.

\subsection{CAL\_HISTORY}

The CAL\_HISTORY sub-table records the calibration context for the
solutions recorded in each main calibration table row. This includes
the parameters used by the calibration solver, the other calibration
tables which may have been applied in the solution, any data
selection, as well as an arbitrary set of notes the user may wish to
attach to the calibration entry for future reference. All CAL\_HISTORY
fields are in string form, with a general keyword-value format.

\definetable{CAL\_MAIN: antenna-based, non-solvable, visibility-plane components}
{
\defline{\bf Columns}
\defline{\em Primary MS indices}
TIME       & Double  &       s  & EPOCH & Interval midpoint \\
TIME\_EXTRA\_PREC & Double & s  &       & Extra TIME precision \\
INTERVAL   & Double  &       s  &       & Calibration interval \\
ANTENNA1   & Int     &          &       & Antenna no.\\
FEED1      & Int     &          &       & Feed on ANTENNA1 \\
FIELD\_ID  & Int     &          &       & Field id.\\
ARRAY\_ID  & Int     &          &       & Subarray number\\
OBSERVATION\_ID & Int  &        &       & Observation id.\\
SCAN\_NUMBER    & Int  &        &       & Scan number  \\ 
PROCESSOR\_ID   & Int  &        &       & Processor id.\\
PHASE\_ID  & Int     &          &       & Phase id.\\
STATE\_ID  & Int     &          &       & State id.\\
PULSAR\_BIN      & Int &        &       & Pulsar bin number \\
PULSAR\_GATE\_ID & Int &        &       & Pulsar gate id. \\   
\defline{\em Secondary MS fields}
FREQ\_GROUP        & Int    &        &       & Freq. group \\
FREQ\_GROUP\_NAME  & String &        &       & Freq. group name \\
FIELD\_NAME        & String &        &       & Field name \\
FIELD\_CODE        & String &        &       & Field code \\
SOURCE\_NAME       & String &        &       & Source name \\
SOURCE\_CODE       & String &        &       & Source code \\
CALIBRATION\_GROUP & Int    &        &       & Source cal. group \\
\defline{\em Gain values}
GAIN               & Complex(a) &    &       & Gain values \\
\defline{\em Reference}
REF\_ANT           & Int(b)     &    &       & Ref. antenna \\
REF\_FEED          & Int(b)     &    &       & Ref. feed \\
REF\_RECEPTOR      & Int(b)     &    &       & Ref. receptor \\
REF\_FREQUENCY     & Double(b)  & Hz & FREQUENCY & Ref. freq. \\
MEAS\_FREQ\_REF    & Int        &    &       & Measures freq. ref.\\
REF\_DIRECTION     & Double(b)  &    & DIRECTION & Ref. dir. \\
MEAS\_DIR\_REF     & Int        &    &       & Measures dir. ref.\\
\defline{\em Sub-table pointers}
CAL\_DESC\_ID      & Int        &    &       & CAL\_DESC index\\
CAL\_HISTORY\_ID   & Int        &    &       & CAL\_HISTORY index\\
}

\begin{description}
\item[Notes:]

\item (i) The dimensions of the array-based columns, which can vary by
row, and are defined in the CAL\_DESC sub-table are as follows: (a)
$(\njones, \njones, \nspw, \nchan)$; and (b) $(\nrecep, \nspw, \nchan)$.

\item (ii) All MS primary indices point into the associated MS, as
defined in the CAL\_DESC sub-table as MS\_NAME. See \aipspp note 229
for a full description of the MS data format.

\item[TIME] Mid-point (not centroid) of calibration interval.

\item[TIME\_EXTRA\_PREC] Extra TIME precision.

\item[INTERVAL] Time interval for which this calibration solution is
valid, or over which it was determined.

\item[ANTENNA1] Antenna number ($\geq 0$), and a direct index into
the ANTENNA sub-table $rownr$ of the associated MS. 

\item[FEED1] Feed number ($\geq 0$).

\item[FIELD\_ID] Field identifier ($\geq 0$).

\item[ARRAY\_ID] Subarray identifier $(\geq 0)$, which identifies data
in separate subarrays.

\item[OBSERVATION\_ID] Observation identifier $(\geq 0)$, which identifies
data from separate observations.

\item[SCAN\_NUMBER] Arbitrary scan number to identify data taken in
the same logical scan. Not required to be unique.

\item[PROCESSOR\_ID] Processor indentifier ($\geq 0$), and a direct
index into the PROCESSOR sub-table $rownr$.

\item[PHASE\_ID] Switching phase identifier ($\geq 0$)

\item[STATE\_ID] State identifier $(\geq 0)$, as defined in Section 3.1.5.

\item[PULSAR\_BIN] Pulsar bin number for the data record. Pulsar data
may be measured for a limited number of pulse phase bins. The pulse
phase bins are described in the PULSAR sub-table and indexed by this
bin number.

\item[PULSAR\_GATE\_ID] Pulsar gate identifier ($\geq 0)$, and a direct
index into the PULSAR\_GATE sub-table $rownr$. 

\item[FREQ\_GROUP] The frequency group to which the spectral window belongs.
This is used to associate spectral windows for joint calibration purposes.

\item[FREQ\_GROUP\_NAME] The frequency group name; user specified.

\item[FIELD\_NAME] Field name; user specified.

\item[FIELD\_CODE] Field code indicating special characteristics of the
field; user specified.

\item[SOURCE\_NAME] Source name; user specified.

\item[SOURCE\_CODE] Source code, used to describe any special characteristics
f the source, such as the nature of a calibrator. Reserved keyword,
including ("BANDPASS CAL").

\item[CALIBRATION\_GROUP] Calibration group number to which this source
belongs; user specified.

\item[GAIN] Array of calibration gain values, expressed as Jones
matrices in an array of dimension (b), which is defined together with
the array coordinates in the CAL\_DESC sub-table.

\item[REF\_ANT] Array of reference antenna numbers, of dimension (b).

\item[REF\_FEED] Array of reference feed numbers, of dimension (b).

\item[REF\_RECEPTOR] Array of reference receptor numbers, of dimension (b).

\item[REF\_FREQUENCY] Array of reference frequencies, of dimension (b).

\item[MEAS\_FREQ\_REF] Array of reference frequency {\it Measure}
references, of dimension (b).

\item[REF\_DIRECTION] Array of reference directions, of dimension (b).

\item[MEAS\_DIR\_REF] Array of reference direction {\it Measure}
references, of dimension (b).

\end{description}


\definetable{CAL\_MAIN: additions for antenna-based, solvable, visibility-plane components}
{
\defline{\bf Columns}
\defline{\em Solution statistics}
TOTAL\_SOLUTION\_OK  & Bool       &    &    & Validity of total solution\\
TOTAL\_FIT           & Float      &    &    & Total fit\\
TOTAL\_FIT\_WEIGHT   & Float      &    &    & Total fit weight\\
SOLUTION\_OK         & Bool(a)    &    &    & Solution validity mask\\
FIT                  & Float(a)   &    &    & Fit array\\
FIT\_WEIGHT          & Float(a)   &    &    & Fit weight array\\
}

\begin{description}
\item[Notes:]

\item (i) The dimension (a) of the array-based columns is: $(\njones,
\njones, \nspw, \nchan)$. These dimensions are defined in the
CAL\_DESC sub-table, and can vary by row.

\item[TOTAL\_SOLUTION\_OK] False if the current calibration solution
interval is flagged, else True.

\item[TOTAL\_FIT] Total fit for the solution interval as a whole, as
$\sqrt{{\chi^2}\over{\Sigma {\rm weight}}}$.

\item[TOTAL\_FIT\_WEIGHT] Total fit weight for the solution interval
as a whole $(\Sigma {\rm weight})$.

\item[SOLUTION\_OK] Validity mask for each element of the calibration
solution array, of dimension (a).

\item[FIT] Fit for each element of the calibration solution array, of
dimension (a), as defined in the TOTAL\_FIT\_WEIGHT description above.

\item[FIT\_WEIGHT] Fit weight for each element of the calibration
solution array, of dimension (a).

\end{description}

\definetable{CAL\_MAIN: additions for interferometer-based, solvable, visibility-plane components}
{
\defline{\bf Columns}
\defline{\em MS primary indices}
ANTENNA2   & Int     &          &       & Second antenna no.\\
FEED2      & Int     &          &       & Feed on ANTENNA2 \\
}

\begin{description}
\item[Notes:]

\item[ANTENNA2] Second antenna number ($\geq 0$) of the
interferometer pair, and a direct index into the ANTENNA sub-table
$rownr$ of the associated MS.

\item[FEED2] Feed number ($\geq 0$) on ANTENNA2.

\end{description}

\definetable{CAL\_DESC: calibration array dimensions and coordinates}
{
\defline{\bf Columns}
NUM\_SPW             & Int             &    &       & No. cal. spw ids\\
NUM\_CHAN            & Int             &    &       & No. cal. freq. chan.\\
NUM\_RECEPTORS       & Int             &    &       & No. cal. receptors\\
N\_JONES             & Int             &    &       & Jones matrix dim.\\
SPECTRAL\_WINDOW\_ID & Int($\nspw$)    &    &       & Associated MS spw ids\\
CHAN\_FREQ       & Double($\nspw,\nchan$) & Hz & FREQUENCY & Cal chan. freq.\\ 
MEAS\_FREQ\_REF  & Int                    &    &    & Measures freq. ref.\\
CHAN\_WIDTH      & Double($\nspw,\nchan$) & Hz &    & Cal. chan. width\\
CHAN\_RANGE      & Int($2,\nspw,\nchan$)  &    &  & Associated MS chan. range\\
POLARIZATION\_TYPE   & String($\nrecep$)  &    &    & Receptor polarization\\
JONES\_TYPE          & String             &    &    & Matrix type\\
MS\_NAME             & String             &    &    & MS name\\
}

\begin{description}

\item[Notes:]

\item[NUM\_SPW] Number of calibration spectral windows.

\item[NUM\_CHAN] Number of calibration frequency channels.

\item[NUM\_RECEPTORS] Number of calibration polarization receptors.

\item[N\_JONES] Dimension of the Jones matrix. For antenna-based
corrections this is less than or equal to the number of polarization
receptors (one or two).

\item[SPECTRAL\_WINDOW\_ID] Associated MS spectral window id.'s. Set
to -1 if the calibration frequency windows span multiple MS spectral
windows.

\item[CHAN\_FREQ] Array of calibration array frequencies.

\item[MEAS\_FREQ\_REF] {\it Measures} reference for the calibration
frequencies.

\item[CHAN\_WIDTH] Array of calibration frequency channel widths.

\item[CHAN\_RANGE] Associated MS frequency channel range.

\item[POLARIZATION\_TYPE] Receptor polarization type (e.g. "R", "L", "X" etc.)

\item[JONES\_TYPE] The matrix type, as a reserved keyword ("DIAGONAL",
"SCALAR", or "GENERAL").

\item[MS\_NAME] Name of the associated MeasurementSet.

\end{description}

\definetable{CAL\_DESC: additions for image-plane components}
{
\defline{\bf Columns}
NUM\_DIR             & Int        &     &                 & No. of directions\\
DIRECTION            & Doube($2,\ndir$) &    & DIRECTION  & Direction array\\
MEAS\_DIR\_REF       & Int              &    &            & Meas. dirn. ref.\\
REGION               & Record($\ndir$)  &    &            & Associated regions.\\
}

\begin{description}

\item[Notes:]

\item (i) Note that the array dimensions in CAL\_MAIN are increased by
$(\ndir)$ for the case of image-plane calibration components.

\item[NUM\_DIR] Number of image-plane calibration directions.

\item[DIRECTION] Array or image-plane calibration directions.

\item[MEAS\_DIR\_REF] {\it Measures} direction reference.

\item[REGION] Associated region for each direction.

\end{description}

\definetable{CAL\_HISTORY: calibration table history}
{
\defline{\bf Columns}
CAL\_PARMS    & String      &    &       & Solver parameters\\
CAL\_TABLES   & String      &    &       & Associated cal. tables\\
CAL\_SELECT   & String      &    &       & Calibration selection\\
CAL\_NOTES    & String      &    &       & Calibration notes\\
}

\begin{description}
\item[Notes:]

\item(i) All CAL\_HISTORY fields are in string form in a general
keyword-value format.

\item[CAL\_PARMS] The parameters used for the solver in deriving the
calibration solution.

\item[CAL\_TABLES] Associated calibration tables applied when solving
for the current Jones component.

\item[CAL\_SELECT] Data selection applied when deriving the
calibration solution.

\item[CAL\_NOTES] Associated user notes for the calibration solution.

\end{description}

\section{References}

\noindent Cornwell, T, 1995, \aipspp Note 183.

\noindent Noordam, J., 1995, \aipspp Note 182.

\noindent Hamaker, Bregman and Sault, 1996, A\&AS, 117, 137.

\end{document}
