%% VLBI++103
%%----------------------------------------------------------------------------
\documentclass[11pt]{article}
\usepackage{html, epsfig, longtable, aips2help}
\input{aips2defs.tex}
%%-----------------------------------------------------------------------------

\begin{document}

\title{Semi-annual report of the NRAO AIPS++ User Group}
  

\author{}

\date{30 Apr, 2001}

\maketitle
%%---------------------------------------------------------------------------
\begin{htmlonly}
\htmladdnormallink{A postscript version of this note is available (76kB).}{../244.ps.gz}
\end{htmlonly}

\tableofcontents
      
\newcommand{\remark}[1]{{\small\begin{itemize}\em \item #1\end{itemize}}}
\newcommand{\prevdocURL}{http://aips2.nrao.edu/aips++/docs/notes/226/226.html}
\newcommand{\thisdocURL}{http://aips2.nrao.edu/aips++/docs/notes/234/234.html}
\newcommand{\systemplanURL}{../220/220.html}
\newcommand{\searchURL}{../../html/aips2search.html}
\newcommand{\aipsviewURL}{../../user/aipsview/aipsview.html}
\newcommand{\GlishmanualURL}{../../reference/Glish/Glish.html}
\newcommand{\GlishtutorialURL}{../../notes/195/195.html}
\newcommand{\synthesisURL}{../../user/synthesis/synthesis.html}
\newcommand{\gsURL}{../../user/gettingstarted/gettingstarted.html}
\newcommand{\userrefmanualURL}{../../user/Refman/Refman.html}
\newcommand{\specsURL}{../../specs/specs.html}
\newcommand{\betaURL}{../beta/beta.html}
\newcommand{\consortiumURL}{../consortium.html}
\externallabels{../../reference/Glish}{../../reference/Glish/labels.pl}
\externallabels{../../user/Utility}{../../user/Utility/labels.pl}
\externallabels{../../user/General}{../../user/General/labels.pl}
\externallabels{../../user/Display}{../../user/Display/labels.pl}
\externallabels{../../user/Dish}{../../user/Dish/labels.pl}
\externallabels{../../user/Synthesis}{../../user/Synthesis/labels.pl}
\externallabels{../../user/ATNF}{../../user/ATNF/labels.pl}
\externallabels{../../user/NFRA}{../../user/NFRA/labels.pl}
\externallabels{../../user/NRAO}{../../user/NRAO/labels.pl}
%

\section{\underbar{Background}}

The past six months have seen significant improvements in aips++ in
several key areas associated with VLA end-to-end processing. 
Detailed
comparisons between data sets processed in a similar fashion in AIPS
and aips++ now show good agreement in the flux scale, calibration
solutions, and final noise levels in the images.  Further, we are now
beginning to process science data that is slated for publication,
rather than just simple test data sets.  However, there is still a
lot
to be done in order to reach the laudable goal of end-to-end VLA
reduction within aips++ for any but the most expert and determined
users.  This goal was recently stated in the April 2001 NRAO
newsletter (see below).  The primary purposes of this document are to
recount the progress made in the past development cycle (for 2001-05,
release 1.5), to summarize the current state and focus of the NRAO
testing, and to draw attention to those areas that we feel most
urgently need work if the aips++ project is going to realize the goal
of usability of the 2001-11 (release 1.6) for VLA continuum data
reduction for general users.          

From the April 2001 NRAO Newsletter (p. 4):

{\it
   "On the synthesis side, many types of VLA observations can now be
   filled, edited, calibrated and imaged entirely within the package.
Our 
   work has concentrated on debugging the path for various types of
   data, adding extra facilities as necessary, and improving the
   documentation. We expect that at the time of the release 1.5  
   [May 2001], the package will be usable by experienced
   interferometrists for VLA reductions, but that novices would be
advised 
   to wait for the subsequent release in the fall of 2001."
}

\section{\underbar{Progress in the Past Cycle}}

We commend the aips++ group for their willingness to assist
members of the NAUG, and for responding to a large number of defects
and enhancement requests.  Since the last release we have seen a
significant number of achievements in aips++ including:

o The ability to fill VLA data using the vlafiller tool.
                                                        
o The implementation (in msplot) of "see it, flag it" style
visibility
  editing.  Some work is still urgently needed in this area (see
below).

o The ability to perform polarization calibration.

o Efforts were made to speed up some parts of aips++ including 
  imaging and the calibrator tool.  Some more work along these lines 
  may still be needed (see below).  

\section{\underbar{Current Status}}

Intensive testing is being carried out in a number of areas, with
emphasis on the path for VLA continuum data.  

o Processing of VLA continuum polarization calibration data (Myers,
Taylor).  The first tests of this path have been successful in
reproducing
AIPS results for C and X band data.  Current roadblocks include the
poor performance of flagging in msplot (the scope of editing in the
X-Y
plot mode, and the lack of support for many AIPS TVFLG functionality
in
the display mode) and lack of CLCOR-like gain-curve and opacity
correction.

o Processing of other VLA continuum data (Fomalont, Taylor, Benson, 
Shepherd) covering all the commonly used VLA continuum bands 
(everything except 4-band and P-band) and including multiple/resolved
sources.  A roadblock here has been the combination of VLA data
from multiple configurations.

Processing of aips++ with spectral-line datasets was a goal
that the NAUG did not achieve in this cycle.  Thus we 
can make no comments or recommendations at this time.  There
will be a concentrated effort in this area over the next 6 months.

\section{\underbar{Immediate Requirements}}

The NAUG identifies the following tasks which must be completed for
aips++ to truly be "usable by experienced interferometrists for VLA
reductions". In order of priority these are:

(1) GUI driven data editing must be fully functional, and work as the
user
  expects.  Since editing will be the first significant task most
users
  encounter within aips++ it is particularly critical that this tool
be
  both intuitive and robust.  You must be able to tell what you are
  editing, and what you are editing must be properly flagged.  This
  includes the ability to undo specific flags (e.g., try out flagging
  some visibilities to see what their effect is).  The missing
  functionality of TVFLG is particularly sorely felt.  The raster
  display in msplot has never worked properly and there is no
capability
  to edit on the basis of rms of amplitude or a running mean.

(2) The user must be able to clean complicated sources, and/or fields
  that include multiple sources.  The ability to define multiple
  fields and multiple clean boxes within each is still missing.

(3) Critical enhancements are still needed for the calibration of
  high-frequency data, including the ability to make corrections
  for elevation dependent gain and opacity effects.

(4) An efficient and inclusive LISTR-like summary output is also 
   urgently needed.

\section{\underbar{Requirements for the next release}}

For the Fall 2001 release (1.6), in order for aips++ to meet its goal
of
being usable by novices, the following would also need to be
achieved:

(1) There must be suitable step-by-step documentation (a.k.a. a
  "cookbook") with supporting scripts that describe the end-to-end
  processing of radio continuum data over the most commonly used
  VLA observing bands (1.3 -> 43 GHz), including polarization. The
  documentation should have minimal aips++ jargon and design
  philosophy, and should concentrate on specific calibration and
imaging
  steps (i.e., what you do, not why you do it), and it should be 
  specifically geared toward users of the VLA, providing frequent 
  referrals to equivalent reduction steps in AIPS. 

(2) Robustness.  At the present time there are still a large number
of 
  bug reports and enhancements coming from the NAUG when bringing new
  datasets through aips++.  All major reduction paths must be tested
  end-to-end against real data without incurring catastrophic
failures
  or critical-path blockages.  The NAUG group is adding data from
  individual members, but we feel a comprehensive "test-plan" might
help
  delineate areas where a wider range of fiducial datasets could be
  obtained.

(3) Aips++ should have computational times that are competitive with
  AIPS, and users should not be unduly slowed down by delays in the 
  GUI interface.  This may require the combination of different
  tools within one execution step.

\section{\underbar{Long-Term Requirements}}

In addition to the core capabilities mentioned above, there are
several enhancements for VLA end-to-end reduction that should see
continued development over the next 6 months:

o Continuum subtraction and bandpass correction for spectral line
data.  

o Wide-field imaging. Concentrate on computing efficiency and ease of
  setting up the multiple fields and clean boxes.  The automatic use
  of the RCP and LCP beam offsets in aips++ is a significant
improvement
  over that in AIPS.  Self-calibration is a must for these fields. 

o Mosaic imaging of continuum and spectral-line data.

o A major overhaul of the documentation.  For example putting links
  under the wrench to take the user to helpful documentation, and
  having help requests that drive a browser to the relevant part 
  of the Getting Results or the Cookbook.  Many of the little 
  pop-up yellow help windows show trivial or obscure comments.
  These should be rationalized and expanded upon.

o The implementation of several Wizards to lead novice users step-by-
  step through VLA calibration and imaging of fairly simple datasets.

o The implementation of a Difmap-like
imaging/self-calibration/editing/
  modelfitting environment.

\end{document}


