%% VLBI++103
%%----------------------------------------------------------------------------
\documentclass[11pt]{article}
\usepackage{html, epsfig, longtable, aips2help}
\input{aips2defs.tex}
%%-----------------------------------------------------------------------------

\begin{document}

\title{AIPS++ DEVELOPMENT PLAN: Release 1.6}
  

\author{Athol Kemball and Tim Cornwell (eds.)\\
   NRAO}

\date{7 May, 2001}

\maketitle
%%---------------------------------------------------------------------------
\begin{htmlonly}
\htmladdnormallink{A postscript version of this note is available (43kB).}{../246.ps.gz}
\end{htmlonly}

\tableofcontents
      
\newcommand{\remark}[1]{{\small\begin{itemize}\em \item #1\end{itemize}}}
\newcommand{\prevdocURL}{http://aips2.nrao.edu/aips++/docs/notes/226/226.html}
\newcommand{\thisdocURL}{http://aips2.nrao.edu/aips++/docs/notes/234/234.html}
\newcommand{\systemplanURL}{../220/220.html}
\newcommand{\searchURL}{../../html/aips2search.html}
\newcommand{\aipsviewURL}{../../user/aipsview/aipsview.html}
\newcommand{\GlishmanualURL}{../../reference/Glish/Glish.html}
\newcommand{\GlishtutorialURL}{../../notes/195/195.html}
\newcommand{\synthesisURL}{../../user/synthesis/synthesis.html}
\newcommand{\gsURL}{../../user/gettingstarted/gettingstarted.html}
\newcommand{\userrefmanualURL}{../../user/Refman/Refman.html}
\newcommand{\specsURL}{../../specs/specs.html}
\newcommand{\betaURL}{../beta/beta.html}
\newcommand{\consortiumURL}{../consortium.html}
\externallabels{../../reference/Glish}{../../reference/Glish/labels.pl}
\externallabels{../../user/Utility}{../../user/Utility/labels.pl}
\externallabels{../../user/General}{../../user/General/labels.pl}
\externallabels{../../user/Display}{../../user/Display/labels.pl}
\externallabels{../../user/Dish}{../../user/Dish/labels.pl}
\externallabels{../../user/Synthesis}{../../user/Synthesis/labels.pl}
\externallabels{../../user/ATNF}{../../user/ATNF/labels.pl}
\externallabels{../../user/NFRA}{../../user/NFRA/labels.pl}
\externallabels{../../user/NRAO}{../../user/NRAO/labels.pl}
%

\clearpage

\section{\underbar{Purpose}}

The purpose of this document is to define the development plan for
AIPS++ release v1.6, currently scheduled for October 15, 2001.

\section{\underbar{Release priorities}}

The highest priority for the project in development cycle v1.6 is to
continue the process of scientific integration. This includes:

\begin{enumerate}

\item Scientific completeness: Demonstrated end-to-end reduction
capabilities for targeted instruments in all scientifically important
observing and reduction modes.

\item Usability by the astronomical community: Quality of the user
interface, applications presentation and user documentation.

\item Robustness, correctness and accuracy: Prove the correctness and
robustness of existing capabilities and ensure comparable performance
to other disk-based packages.

\item Continued deployment to an expanded user base: Continue to
increase the scientific user base for AIPS++.

\end{enumerate}

Maintaining a sound infrastructure in the project is important
however, and infrastructure work has been scheduled for this cycle
carefully; primarily in areas which are vital for the long-term
vitality of the project, or which are in the critical path for
application development.

\section{\underbar{Introduction}}

Development targets are listed separately for each major development
area in AIPS++. These targets cover only the development cycle through
October 2001, and do not include longer-term items, which are tracked
separately for consideration in subsequent development cycles (and can
be found in docs/project/deferred\_1.6.ps.gz). There are 21 weeks in
this development cycle, between the week of May 7, 2001 and the
expected code-freeze on 1 October 2001, but deductions for defect
correction and code review (30\%; 1.5 days per week), user support (up
to 20\%), and where applicable, science time (25\%), have been
made. The exact numbers vary by developer and institution, and the
best estimate has been made in each case. Variability in these
estimates is accommodated by the inclusion of medium prioirty targets,
which may need to be carried over into the next cycle. Each target is
assigned a priority as high ({\bf H}), which implies that the target
is expected to be completed this cycle, and medium ({\bf M}), which is
work to be done on an as-available basis. Note also, that not all
developers are available to the project on a full-time basis, and have
reduced commitments as a result.

\vskip 0.5cm

\begin{tabular}{|l|l|l|}\hline
{\bf Abbr.}&{\bf Developer}&{\bf Affiliation}\\ \hline\hline
AK & Athol Kemball & NRAO\\
AS & Anuj Sarma & NCSA\\
BG & Bob Garwood & NRAO\\
DK & David King & NRAO\\
DM & Dave Mehringer & NCSA\\
DrS & Darrell Schiebel & NRAO\\
GvD & Ger van Diepen & ASTRON\\
GM & George Moellenbrock & NRAO\\
HR & Harold Ravlin & NCSA\\
JB & Jim Braatz & NRAO\\
JN & Jan Noordam & ASTRON\\
JM & Joe McMullin & NRAO\\
KG & Kumar Golap & NRAO\\
MM & Malte Marquarding & ATNF \\
MW & Mark Wieringa & ATNF \\
NK & Neil Killeen & ATNF\\
PC & Paulo Cortes & NCSA\\
PT & Peter Teuben & BIMA \\
RM & Ralph Marson & NRAO\\
RP & Ray Plante & NCSA\\
OS & Oleg Smirnov & ASTRON\\
TC & Tim Cornwell & NRAO\\
WB & Wim Brouw & ATNF\\
WY & Wes Young & NRAO\\
\hline
\end{tabular}

%%----------------------------------------------------------------------

\section{\underbar{Single-dish}}



\begin{enumerate}

\item Complete GBT MS filler: Including: i) complete IF/LO conversion;
ii) GBT GO FITS file to MS; iii) Tcal to SYSCAL sub-table; iv)
spectrometer and spectral processor backend support; v) track M\&C FITS
file changes; vi) improve reliability and robustness in on-line mode;
vii) complete holography support; viii) document filler internals in
preparation for filler transfer to GBT science support. ({\bf BG, H, 9
wk}).

\item SD MS audit: Audit the MS produced by the GBT MS filler for
scientific completeness. ({\bf BG, H, 2 wk, AK, H, 1 wk}).

\item SD calibration: Complete basic SD calibration modes using
synthesis infrastructure({\bf JM, H, 6 wk, AK, H, 2 wk}).

\item ssa enhancements: Complete ssa improvements. ({\bf JM, H, 2 wk}).

\item IARDS: Interim real-time display system for the GBT ({\bf JB, H, 4 wk}).

\item Holography support: Full holography support using synthesis
infrastructure. ({\bf JB, H, 4 wk}).

\item Document DCR tool: Document DCR tool and heuristics({\bf JB, H, 3 wk}).

\item SD imaging: Including: i) verify single-dish imaging speed for
spectral line data; ii) demonstrate combined VLA and GBT
imaging. ({\bf TC, H, 6 wk}).

\end{enumerate}

%%----------------------------------------------------------------------

\section{\underbar{Synthesis}}



\begin{enumerate}

\item Complete the FITS-IDI filler: Including: i) support for all
required MS sub-tables; ii) VLBA correlator quantization corrections;
iii) testing with SMA and VLBA data. Work with SMA in these
tasks. ({\bf GM, H, 2 wk}).

\item BIMA filler revisions: Continued improvements in the BIMA
filler. ({\bf RP, H, 1 wk}).

\item Multi-MS support in imager and calibrater: Make existing
multi-MS support available in the calibrater and imager tools. ({\bf
KG, H, 3 wk}).

\item Support for variable spw: Ensure proper functioning of the
synthesis tools with variable-shaped spectral windows. Includes
resolution of all current defects in this area ({\bf KG, H, 2 wk}).

\item Implement initial compressed data format: Implement initial
solution to the MS data volume problem caused by the MODEL\_DATA,
CORRECTED\_DATA and IMAGING\_WEIGHT scratch columns. ({\bf GvD, H, 2
wk, \bf AK, H, 1 wk}).

\item msplot defect correction: Fix all current, high-severity msplot
defects. ({\bf RM, H, 2 wk}).

\item Autoflag interface revisions: Change autoflag.g Glish interface
to be compatible with AIPS++ user interface conventions. ({\bf OS, H,
1 wk}).

\item Autoflag C++ code review: Complete preparation of autoflag C++
code for AIPS++ code review. ({\bf OS, H, 2 wk, \bf AK, H, 1 wk}).

\item Requirements for uv-component fitting: Finalize requirements and
implementation document for full uv-component fitting. Including list
of required changes to component models. ({\bf WB, H, 1 wk}).

\item Analytic errors in simulator: Add basic analytic errors to
simulator, including: i) Chebyshev polynomial in amplitude, phase,
rate and delay (G Jones); ii) polynomial bandpass (B Jones); iii)
Chebyshev polynomial in instrumental polarization (D Jones). Requires
Chebyshev polynomial implementation. ({\bf RP, H, 2 wk}).

\item calibrater enhancements: Includes: i) identified improvements in
the MS access classes; ii) full interpolation and parametrization
modes. ({\bf GM, H, 4 wk}).

\item Ionosphere C++ code review: Prepare Ionosphere C++ code for code
review. ({\bf OS, H, 2 wk}).

\item Ionsophere correction testing: Test first-order FJones
correction on WSRT and other data. ({\bf OS, H, 3 wk}).

\item Second-order ionosphere corrections: Derive initial second-order
FJones corrections to the PIM model. ({\bf OS, H, 6 wk}).

\item Automated wide-field imaging: Refine the packaging of the
existing wide-field imaging capabilities. Includes: i) improved mask
setting, including automated options; ii) improvements to
dragon.g. ({\bf KG, H, 3 wk}).

\item SD mosaic combinations: Improved single-dish mosaicing data
combination options. ({\bf KG, H, 3 wk}).

\item Planet flux density calibration: Develop flux density calibration
using planets. ({\bf RP, H, 2 wk}).

\item imager test scripts: Revise all imager test scripts to use
simulated data for test generation and verification. Includes coverage
of all observing contexts (single-field, multi-field, mosaiced, and
wide-field), and all available deconvolution algorithms. ({\bf DM, H,
4 wk, KG, H, 1 wk}).

\item ATCA test scripts: Test scripts and associated data
demonstrating ATCA reduction for continuum, continuum polarimetry and
spectral line reduction modes. ({\bf MW, H, 3 wk}).

\item WSRT test scripts: Test scripts and associated data
demonstrating WSRT reduction for continuum, continuum polarimetry and
spectral line reduction modes. ({\bf JN, H, 4 wk}).

\item BIMA test scripts: Additions to the existing BIMA test scripts
to cover new observing modes. ({\bf DM, H, 2 wk}).

\item VLA test scripts: Expand the existing VLA test scripts to cover spectral-line modes and additional data sets. ({\bf AK, H, 1 wk}).

\item IRAM AIPS++ test: Verify AIPS++ reduction using IRAM test
data. ({\bf AK, H, 5 wk, KG, M, 2 wk}).

\end{enumerate}

%%----------------------------------------------------------------------

\section{\underbar{Documentation}}



\begin{enumerate}

\item Editing of Getting Results: Includes: i) edit existing and new
GR chapters to conform to a common framework; ii) identify missing
areas in GR; iii) incorporate user group feedback in the editing
process. ({\bf NK, H, 4 wk}).

\item Single-dish documentation editing: Improve the existing SD
documentation and add new material ({\bf JM, H, 2 wk}).

\item ATCA Getting Results chapter: Getting Results chapter for ATCA
continuum, continuum polarimetry and spectral line reduction
modes. ({\bf MW, H, 2 wk}).

\item BIMA Getting Results chapter: Getting Results chapter for BIMA
continuum, continuum polarimetry and spectral line reduction
modes. ({\bf AS, H, 2 wk}).

\item VLA Getting Results chapter: Edit and extend the existing VLA
Getting Results chapter to cover spectral line, and improve existing
material. ({\bf GM, H, 2 wk}).

\item Recipes: Revitalize the existing recipes repository. Change the
format as required, and add new recipes in all areas. ({\bf GM, H, 3
wk}).

\end{enumerate}

%%----------------------------------------------------------------------

\section{\underbar{Applications integration}}

This includes development aimed at improving the overall integration
of applications in AIPS++ in a common framework, using common
interfaces and services.



\begin{enumerate}

\item History propagation: Develop infrastructure for history
propagation. ({\bf RP, H, 3 wk}).

\item Programmer documentation: Update the existing programmer
documentation. ({\bf AK, H, 1 wk}).

\end{enumerate}

%%----------------------------------------------------------------------

\section{\underbar{Build and code distribution system}}

This covers all system work, excluding maintenance and support of the
basic library.



\begin{enumerate}

\item Documentation system: Develop written requirements for the
documentation build system; audit current capabilities. Explore
central documentation build and remote distribution using cvsup ({\bf
WY, H, 2 wk}).

\item Glish documentation utilities: Develop utilities to assist in
generating Glish-level tool documentation, and to verify completeness
and correctness of existing Glish documentation. ({\bf WY, H, 3 wk}).

\item Build and release maintenance: Maintenance of existing builds
under Linux, Solaris and IRIX. Patch generation and distribution for
the release master. ({\bf WY, H, 4 wk}).

\item Solaris native C++ port: Complete port of AIPS++ to the Solaris
native C++ compiler. Document all required changes as a note. ({\bf
WY, H, 2 wk}).

\end{enumerate}

%%----------------------------------------------------------------------

\section{\underbar{Glish}}



\begin{enumerate}

\item GlishTk performance: Continue evaluation of GlishTk performance
factors, including: i) dependence on pipe buffer size; ii) other
short-term optimizations to mitigate dynamic scheduling losses. ({\bf
DrS, H, 3 wk}).

\item Glish threading: Assess cost of Glish threading. Prototype
threading in parts of the Glish library and determine likely GTk
performance enhancements. ({\bf DrS, H, 3 wk}).

\item fail on return: Implement fail on return handling in Glish at
function end-points. ({\bf DrS, H, 1 wk}).

\item Defect backlog: Address current defect backlog, including: i)
slider problems; ii) hourglass packing; iii) remaining JIVE reference
counting defects. ({\bf DrS, H, 3 wk}).

\item Tab widget: Add the Tab widget. ({\bf DrS, H, 2 wk}).

\end{enumerate}

%%----------------------------------------------------------------------

\section{\underbar{User interface and tasking}}



\begin{enumerate}

\item GUI performance improvements: Improve GUI performance by
investigating the use of larger-scale widgets, and/or optimization of
existing Glish GUI code. ({\bf RM, H, 3 wk}).

\item Data item propagation: Ensure use of all existing data items
throughout the user interface, where possible. ({\bf RM, H, 2 wk}).

\item guientry defects: Resolve all current guientry.g defects. ({\bf
RM, H, 2 wk}).

\item toolmanager enhancements: Includes: i) allow aipsrc
customization of tools visible in toolmanager; ii) trial
implementation of default sets for one tool. ({\bf RM, H, 2 wk}).

\item Revised tasking requirements: Develop requirements for an
improved Tasking system. ({\bf AK, H, 1 wk}).

\item CORBA evaluation: Develop implementation plan for CORBA adoption
in the Tasking system and Glish. Includes some CORBA
prototyping. ({\bf DrS, H, 3 wk}).

\end{enumerate}

%%----------------------------------------------------------------------

\section{\underbar{Basic library}}



\begin{enumerate}

\item Overhaul Functionals: Overhaul Functionals so that they can be
used more extensively in other parts of the library, including
Fitting. Includes an implementation of optimized
auto-differentiation. ({\bf WB, H, 4 wk}).

\item C++ standard proposal: Draft a proposal outlining strategies for
implementing further C++ standard library compliance in AIPS++. ({\bf
WB, H, 1 wk}).

\item C++ interface to non-linear fitting: Improve the C++ interface
to existing non-linear fitting capabilities. Requires improvements in
Functionals ({\bf WB, H, 2 wk}).

\item Deep copy: Allow deep copy of tables. ({\bf GvD, H, 1 wk}).

\item Measures frame handling: Improved frame handling in
Measures. ({\bf WB, M, 1 wk}).

\end{enumerate}

%%----------------------------------------------------------------------

\section{\underbar{Images}}



\begin{enumerate}

\item LEL spectral index function: Implement a spectral index function
in LEL ({\bf GvD, H, 0.5 wk}).

\item Support non-cardinal axes: Support slicing along non-cardinal
axes, which is required for the pvslice visualization
application. Requires addition of Lattice interpolation along
non-cardinal axes. ({\bf NK, H, 2 wk, \bf GvD, H, 3 wk}).

\item Mask generation from cubes: Support mask generation from cubes,
as required to support a range of visualization and image analysis
applications. ({\bf NK, H, 3 wk}).

\item Multi-dimensional LEL: Add facility for handling multi-dimensional images in LEL. {(\bf GvD, H, 1 wk}).

\item Unified logsink use in Image classes: Unify logsink use in
derived Image classes. ({\bf GvD, H, 1 wk}).

\item Miriad image handling: Support MIRIAD image handling. ({\bf PT,
M, 4 wk}).


\end{enumerate}

%%----------------------------------------------------------------------

\section{\underbar{Visualization}}

This includes work in the Display Library (DL), and in applications
using this library.



\begin{enumerate}

\item pvslice application: Implement high-level application supporting
pv-slicing. ({\bf MM, H, 4 wk}).

\item Color bar wedge: Implement color bar wedge support in the DL and
viewer. ({\bf MM, H, 2 wk}).

\item Multiple regions per panel: Support multiple regions per display
panel. ({\bf MM, H, 1 wk}).

\item Unify CachingDD and PADD code branches: Unify the existing
caching and principal axes DisplayData code branches. Preserve the
current API for each branch in the interim to allow uv DD development
to continue. ({\bf MM, H, 5 wk}).

\item Proposal for coordinates on world canvas: Prepare a
proposal/design for the implementation of coordinates on the
WorldCanvas. ({\bf MM, H, 2 wk}).

\item uv DisplayDatas: Complete: i) MSAsRaster; ii) MSAsContour; iii)
MSAsXY; iv) TableAsRaster; v) TableAsContour; vi) TableAsXY. ({\bf DK,
H, 6 wk}).

\item Integrated uv-visualization: Provide full integration of the
uv-DD through to the viewer, and make available as re-usable
uv-visualization components. ({\bf DK, H, 4 wk}).

\item Editing event-handlers: Develop editing capabilities as DL event
handlers. ({\bf DK, H, 4 wk}).

\item Animation increments in world coordinates: Allow animation
increments in world coordinates. ({\bf MM, M, 1 wk}).

\item OpenGL PixelCanvas: Continue development of an OpenGL or vtk
PixelCanvas. ({\bf HR, M, 4 wk}).

\end{enumerate}

%%----------------------------------------------------------------------


\section{\underbar{Parallelization and high-performance computing}}


\begin{enumerate}

\item Revise test scripts for parallel problems: Revise the test
scripts for large parallelized algorithms to use simulated test data
and verification. ({\bf DM, H, 2 wk}).

\item Key science processing and profiling: Reduce the identified key
science datasets on DSM and cluster systems, with and without
associated performance and I/O profiling. ({\bf AS, H, 6 wk}).

\item IA64 port: AIPS++ port to the IA-64 Linux cluster. ({\bf PC, H,
4 wk}).

\item Parallel I/O: Initial implementation of parallel I/O
capabilities. ({\bf WY, H, 3 wk}).

\end{enumerate}    

%%----------------------------------------------------------------------

\end{document}


