\documentclass[11pt]{article}
\usepackage{html, epsfig, longtable, aips2help}
\input{aips2defs.tex}
%%-----------------------------------------------------------------------------

\begin{document}

\title{Updating AIPS++}


\author{Wesley Young\\
   NRAO}

   \date{7 Nov, 2002}

   \maketitle
   %%---------------------------------------------------------------------------
   \begin{htmlonly}
   \htmladdnormallink{A postscript version of this note is available (28kB).}{../252.ps.gz}
   \end{htmlonly}

   \tableofcontents
\section{Introduction}

The current mode of applying patches to AIPS++ is spotty at best for addressing user requests for defect
resolutions and enhancements.  More frequent patches would be helpful but they are difficult to apply
and are often
precluded by the project's desire to only apply bug fixes to release.  A possibly more useful model is 
to allow users access to the project's stable builds. 

\section{Non-consortium User Updates}

In our current scheme, consortium sites and those sites that build from source have access to daily
changes in the code tree.  Available to these sites, is a stable build mechanism inhale -b.  Users who
install from CD don't readily have this option (although linux users can use the inhale binary option
provided they make adjustments to their systems).

\subsection{A New Update Scheme}
\begin{description}
\item[stable - base] Our current stable. 
\item[stable - updated] An updated stable with 
new features, enhancements, or bug fixes, stuff is plugged in but must pass
all stable tests to be moved into updated stable tree.
\item[beta]  First stable after slushy code freeze
\item[released]  Beta that meets all release criteria.

\item [additional explanation]  We continue to test for stables in our current mode, this produces our
    stable-base.  The stable-base is copied to stable-updated.  Stable-updated contains apps
    and scripts that are tested to work with the current stable but are not in the current
    stable.  Essentially we replace the app and/or scripts with new ones and run an assay on
    the "updated stable" to make sure it still passes, if so the stable is then updated.
\end{description}

\subsection{ISO images}
We'll have ISO release images 
available for base stables, betas, and releases.  This will require us 
to move these ISO images to ATNF, UIUC, and ASTRON.

\subsection{Use cvs/cvsup}
Put stable bin files (base and updated), libexec files, and shared libraries under cvs/cvsup.

\subsection{Create a new script, aupdate}
\begin{description}
\item[Options]
\item[stable] - get updated stable version, with a noupdate option to get only the base stable.
\item[beta] - get beta version
\item[release] -only get released version (includes patches)
\item[(no)bin] - get updated bin files
\item[(no)libexec] - get updated libexec files
\item[(no)lib] - get updated shared libraries
\item[(no)docs] - get  updated docs
\item[(no)basedata] - get  updated base data
\item[(no)auxdata] - get  updated aux data (not aux could be telescope specific)
\item[(no)code] - get updated code tree
\item[usefile(s)] - update from a local tar file.
\item[default] - is stable binary libexec lib basedata docs
\item[silent (-q)] - Just do it don't ask for confirmation.
\end{description}

Aupdate would normally use cvsup to update various directory trees. The usefile option would allow 
downloading a tar file and updating the system for those machines that don't support cvsup.


\subsection{What the user does}

\begin{enumerate}
\item Install release
\item Choose one of three update options
   \begin{enumerate}
   \item To update to the latest alpha release (stable):
   aupdate
   \item To update to the lastest beta release (post slushy freeze):
   aupdate beta
   \item To get the latest release (including patches)
   aupdate release
   \end{enumerate}

\item aupdate will tell the user something like

\begin{verbatim}
    $aupdate
    NOTICE - NOTICE - NOTICE

    AIPS++ binary files tend to be big, you should really be on a
    decent network link to down load the updates

    AIPS++update ->
       Updating the following linux/bin linux/lib linux/libexec docs from stable
    Is this correct?

    If y then do the update, else if n then
    AIPS++update ->
       Please specify stable|beta|release (no)bin (no)lib (no)libexec (no)docs
       (no)code (no)basedata
    update?

    Repeat until we get a y.  Using a -q or silent avoids all these questions,
    and figures out what to do from the installation.
\end{verbatim}
\end{enumerate}

\subsection{Notes}

\begin{enumerate}
\item An alternative would be for the user to download the update files and 
use the usefile(s) option.

\item The user who owns the AIPS++ installation must run the aupdate script.
\end{enumerate}
\section{Action Items}
\begin{enumerate}
\item Create the stable - update tree.
\item Put stable base/update under cvs/cvsup.
\item Write the aupdate script.
\item Create the tar files for systems that don't support cvsup.
\item Set the test procedures for adding files/apps to the updated stable.
\item Generate additional ISO images for the base stable.
\end{enumerate}


\end{document}
